---
title: "Backtest: holding out the latest diagonals to validate projections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backtest: holding out the latest diagonals to validate projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

## Motivation

Reserving and projection methods are fitted on observed data, but their
practical value lies in how they would have performed at past valuation
dates. `backtest()` answers that question by hiding the latest `holdout`
calendar diagonals from a triangle, refitting the model on the earlier
portion, and comparing its projection to the actuals that were withheld.
This is calendar-diagonal hold-out (rather than dev-period hold-out),
because it simulates "what would the model have said *K* months ago at
the valuation date?". The cell-level metric (`ae_err`, "A/E Error")
follows the standard actuarial A/E convention,
$\mathrm{ae\_err} = v_{\mathrm{actual}} / v_{\mathrm{pred}} - 1$, where
positive values flag under-projection (the model under-estimated;
actual exceeded expected) and negative values flag over-projection.

## Basic usage

```{r}
library(lossratio)
data(experience)
tri_sur <- build_triangle(
  experience[coverage == "SUR"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "loss_incr",
  premium  = "premium_incr"
)

bt <- backtest(tri_sur, holdout = 6L)
print(bt)
```

The returned object is a `"Backtest"` list with these key slots:

- `ae_err` — per-cell `data.table` (cohort, dev, actual,
  expected, aeg, ae_err + `_incr` siblings, calendar_idx).
- `col_summary` — A/E Error aggregated by `dev`.
- `diag_summary` — A/E Error aggregated by calendar diagonal.
- `masked` — the triangle the fit was trained on (latest diagonals removed).
- `fit` — the fit object returned by the target-specific dispatcher
  (`fit_lr` / `fit_loss` / `fit_premium`) chosen by `target=`.

`summary(bt)` prints the two summary tables alongside the call metadata.

## Validation coverage after masking

Masking the latest `holdout` diagonals shortens the triangle's
lower-right edge. Chain ladder can only project as far as the largest
dev still observed in the masked data, so cells beyond that range — the
oldest cohorts at their latest dev — have no projection to compare
against. These unreachable cells are silently dropped, so `bt$ae_err`
contains only cells where both an actual and a finite projection exist.

Practical takeaway: as `holdout` grows, the validation set shrinks
fastest in the oldest cohorts' late-dev region — exactly where chain
ladder relies on extrapolation (projection beyond the observed dev
range), so it is the area most in need of validation yet the first to
disappear.

## Output interpretation

**`col_summary` — systematic bias by development period.** A
consistently signed A/E Error at a given dev signals a structural
mismatch between the model and that maturity. Early-dev positive
values usually reflect inflated link factors; late-dev values flag
tail miscalibration.

```{r}
head(bt$col_summary, 8)
```

`ae_err_mean` averages cell-level A/E Error, `ae_err_med` is the
median, and `ae_err_wt = sum(actual - proj) / sum(proj)` is the
exposure-weighted pooled A/E ratio minus 1. Comparing the three
columns flags whether a few large cells dominate (`ae_err_wt` very
different from `ae_err_med`) or the bias is uniform.

**`diag_summary` — calendar-year effect.** A single bad diagonal in
otherwise unbiased output points at a calendar event (a rate change,
claim handling shift, or one-off shock) that a static fitter cannot
see by construction.

```{r}
bt$diag_summary
```

A monotone drift across calendar diagonals (as in the SUR example
above, where A/E Error becomes increasingly positive across
`25, ..., 30`) typically indicates that actuals on the latest
diagonals are running above what the earlier-cohort link factors
imply, i.e. a regime shift the static model has not absorbed.

**`ae_err` — cell-level outliers.** For diagnosing specific
cohort × dev cells, inspect `bt$ae_err` directly:

```{r}
head(bt$ae_err, 5)
```

## Plot demos

Four plot views are registered on `"Backtest"`:

```{r}
plot(bt, type = "col")    # A/E Error by dev (point + dashed zero line)
plot(bt, type = "diag")   # A/E Error by calendar diagonal
plot(bt, type = "cell")   # per-cohort A/E Error trajectories over dev
plot_triangle(bt)         # diverging-color heatmap on the held-out wedge
```

`type = "col"` is the right place to look for systematic dev-period
bias; `type = "diag"` reveals calendar-year drift; `type = "cell"`
exposes which cohorts contribute the bias; `plot_triangle()` puts the
cell-level A/E Error values on the same triangular layout as
`plot_triangle()` for the underlying fit, with a red/blue diverging
palette where red marks under-projection (actual > pred) and blue
marks over-projection (actual < pred).

## Holdout selection

Choose `holdout` to balance two opposing effects:

- Too large: the masked triangle loses its latest experience, so the
  oldest cohorts have few or no reachable cells in their later dev
  periods. The validation set shrinks unevenly, biased toward early
  dev.
- Too small: the held-out wedge is just a thin diagonal band, which
  may not capture enough cells to reveal systematic patterns.

Typical choices are `holdout = 6L` (half-year) for monthly triangles,
or `holdout = 12L` (full year) for stronger validation when the
triangle has at least 24–30 diagonals of history.

## Choosing the projection target

The default is `target = "lr"` with `loss_method = "sa"`. The loss
ratio is unitless and dimension-free across cohorts of very different
volume, so `ae_err_mean` and `ae_err_med` carry a consistent meaning
across the triangle.

> **A note on `target`.** `target` is the **score column** — the column
> on which actual vs. predicted are compared cell-by-cell. It selects
> which role-specific fitter `backtest()` runs internally and which
> projection column on the fit's `$full` table is compared against the
> held-out actuals:

| `target`    | Internal fitter      | Method arg       | Compared column |
|-------------|----------------------|------------------|-----------------|
| `"lr"`      | `fit_lr()`           | `loss_method`    | `lr_proj`       |
| `"loss"`    | `fit_loss()`         | `loss_method`    | `loss_proj`     |
| `"premium"` | `fit_premium()`      | `premium_method` | `premium_proj`  |

The `loss_method` argument selects the underlying loss / loss-ratio
projection strategy: `"sa"` (stage-adaptive, the default) blends
exposure-driven projections before the maturity point with chain
ladder afterwards; `"ed"` is purely exposure-driven; `"cl"` is the
classical chain ladder. The `premium_method` argument selects the
premium projection strategy when `target = "premium"`.

```{r}
bt_sa       <- backtest(tri_sur, holdout = 6L, loss_method = "sa")  # default
bt_ed       <- backtest(tri_sur, holdout = 6L, loss_method = "ed")
bt_cl       <- backtest(tri_sur, holdout = 6L, loss_method = "cl")

bt_loss     <- backtest(tri_sur, holdout = 6L,
                        target = "loss", loss_method = "cl")
bt_premium  <- backtest(tri_sur, holdout = 6L,
                        target = "premium", premium_method = "cl")

print(bt_sa)
```

For monetary impact (loss or premium) backtesting, set
`target = "loss"` or `target = "premium"` to score the corresponding
projection lane directly.

## See also

- `vignette("chain-ladder-reserving")` — `fit_cl()` reference.
- `vignette("projection")` — `fit_lr()` and the `"sa"`,
  `"ed"`, `"cl"` methods.
- `?backtest`, `?plot.Backtest`, `?plot_triangle.Backtest`.
