---
title: "(Reference) Chain ladder reserving"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{(Reference) Chain ladder reserving}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

> **Reference: P&C reserving context.** This article covers chain
> ladder *reserving* — projecting ultimate paid / incurred loss for an
> open accident year — which is a Property & Casualty (P&C, 손해보험)
> use case. The lossratio package's primary focus is long-term health
> insurance loss ratio projection (`fit_lr`), where the reserving
> framing applies only loosely. We include this article so practitioners
> coming from a P&C background see how `fit_cl` maps to the classical
> Mack chain ladder workflow they're already familiar with.

`fit_cl()` is the dedicated chain ladder fit for a single value
column. Unlike `fit_lr()` (which projects loss / exposure jointly to
get loss ratio), `fit_cl()` projects one cumulative metric forward
and computes Mack-style standard errors per cohort.

## Basic usage

For brevity this vignette uses the `SUR` group only — every step
generalises to multi-group input.

```{r}
library(lossratio)
data(experience)
tri <- as_triangle(
  experience[coverage == "SUR"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_prem"
)

cl <- fit_cl(tri, target = "loss", method = "mack")
print(cl)
```

`target` selects the cumulative column to project — typically
`"loss"` (cumulative loss) for reserving, or `"premium"` (cumulative
risk premium) for exposure projection.

## Mack chain ladder

`fit_cl()` implements the Mack (1993) chain ladder. Adjacent
development links are summarised by age-to-age factors
$f_k = C^L_{k+1} / C^L_k$ — selected per link and then chained
to project each cohort forward to ultimate. On top of the point
projection, Mack's formulae decompose the prediction variance
into process and parameter components, yielding per-cell
standard errors and confidence intervals.

```{r}
cl_mack <- fit_cl(tri, target = "loss", method = "mack")

# $full and $summary carry both the projection and its variance
head(cl_mack$summary)
```

The projection plot's confidence bands (`show_interval = TRUE`)
use those variance estimates:

```{r}
plot(cl_mack, type = "projection", show_interval = TRUE)
```

## Tail factor

For triangles where the latest observed development period is still
developing, an extrapolated tail factor estimates ultimate:

```{r}
# Log-linear extrapolation from the selected ATA factors
cl_tail <- fit_cl(tri, target = "loss", method = "mack", tail = TRUE)

# Or supply a literal tail factor
cl_tail <- fit_cl(tri, target = "loss", method = "mack", tail = 1.025)
```

The extrapolation fits $\log(f_k - 1) \sim k$ to projected factors and
extends the projection by the cumulative product of extrapolated
$f_k$ values. Disabled by default (`tail = FALSE`).

## Maturity filtering

If selected ATA factors are volatile, restrict projection to the
mature region only:

```{r}
cl_mat <- fit_cl(
  tri,
  target   = "loss",
  method   = "mack",
  maturity = maturity_spec(max_cv = 0.10, max_rse = 0.05)
)

cl_mat$maturity
```

`maturity_spec(...)` captures custom `detect_maturity()` thresholds
and is invoked internally on the (possibly masked) triangle. Pass
`maturity = "auto"` for default thresholds, a pre-built `Maturity`
object for a fixed override, or `maturity_at(...)` for a manual
per-group $k^*$.

## Variance components (Mack)

`fit_cl(method = "mack")` decomposes the projection variance into:

- `target_proc_se` — process variance, from $\sigma^2_k$ (residual link
  variance per development period).
- `target_param_se` — parameter variance, from the uncertainty of the
  selected age-to-age factors $\hat{f}_k$.
- `target_total_se` — total standard error,
  $\sqrt{\mathrm{target\_proc\_se}^2 + \mathrm{target\_param\_se}^2}$.
- `target_total_cv` — coefficient of variation,
  `target_total_se / target_proj`.

```{r}
summary(cl_mack)
```

## Reserve plot

`type = "reserve"` shows reserve per cohort with optional error bars
(Mack only):

```{r}
plot(cl_mack, type = "reserve", conf_level = 0.95)
```

## Triangle visualisation

`plot_triangle()` displays the cohort × dev cells as a heatmap,
distinguishing observed cells from projected:

```{r}
plot_triangle(cl_mack, region = "full")    # observed + projected
plot_triangle(cl_mack, region = "proj")    # projected only
plot_triangle(cl_mack, region = "data")    # observed only
```

The `label_style = "cv"` mode shows coefficient of variation per
cell, useful for spotting unreliable cells:

```{r}
plot_triangle(cl_mack, label_style = "cv")
plot_triangle(cl_mack, label_style = "se")
plot_triangle(cl_mack, label_style = "ci")
```

## Sigma extrapolation methods

Mack variance requires $\sigma_k$ at all development links, including
the last where it cannot be estimated directly. `sigma_method`
controls the extrapolation:

| `sigma_method`  | Behaviour                                                             |
|-----------------|-----------------------------------------------------------------------|
| `"locf"`        | (default) last observation carried forward                            |
| `"min_last2"`   | min of the last two estimable $\sigma$ values — conservative          |
| `"loglinear"`   | Log-linear extrapolation from the observed $\sigma_k$ sequence        |

```{r}
fit_cl(tri, target = "loss", method = "mack", sigma_method = "loglinear")
```

## See also

- `vignette("projection")` — when to use `fit_lr()` instead.
- `vignette("triangle-link-and-maturity")` — `summary()`,
  `detect_maturity()`, ata diagnostic plots.
- `?fit_cl`, `?detect_maturity`, `?fit_ata`.
