---
title: "Convergence: detecting when projected loss ratio stops revising"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence: detecting when projected loss ratio stops revising}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

## Motivation

`detect_maturity()` answers the question *"from which development
period are link factors $f_k$ reproducible across cohorts?"*. That is
necessary for chain-ladder projection but not sufficient for declaring
a portfolio's projected loss ratio converged: in long-duration health
insurance both $f_k \to 1$ and $g_k \to 0$ arise mechanically from
cumulative denominators growing, regardless of whether the underlying
experience has actually settled. A criterion built on those quantities
passes automatically with $k$, not because of true convergence — what
we have called the *inertia* failure mode.

`detect_convergence()` detects the **convergence point** $k^{**}$ —
the first valuation $v \ge k^*$ at which the projected loss ratio has
predictively converged. It is the natural counterpart to $k^*$
(maturity point, from `detect_maturity()`): $k^*$ marks where link
factors $f_k$ become reproducible, while $k^{**}$ marks where the
projection itself stops moving with new data. Long-duration health
portfolios may cross $k^*$ early yet remain far from $k^{**}$.

The detector combines two orthogonal conditions, both required to
hold for $M$ consecutive valuations:

1. **Predictive revision** is small relative to its parameter SE:
   $R_v < c \cdot \hat{SE}^{\mathrm{param}}_v$, where
   $R_v = |\hat{LR}^{\mathrm{proj}}_v(D_v) -
   \hat{LR}^{\mathrm{proj}}_v(D_{v-1})|$ is the change in the
   projected portfolio LR caused by adding one new calendar diagonal.
2. **Cross-cohort dispersion** of incremental LR is small:
   $\hat{D}_v < \tau$, where
   $\hat{D}_v = 1.4826 \cdot \mathrm{MAD}_i(\hat{lr}_{i,v}) /
   |\mathrm{median}_i(\hat{lr}_{i,v})|$.

Operating $\hat{D}_v$ on **incremental** rather than cumulative
loss ratio keeps it inertia-free — per-period quantities have no
cumulative denominator to dampen them. The two clauses guard against
complementary failure modes: $R_v$ checks that the *model output* has
stopped revising, while $\hat{D}_v$ checks that the *raw
period-by-period experience* is genuinely consistent across cohorts
at that dev. Either alone can be fooled — in chain-ladder projection
the mechanical drift $\hat{f}_k \to 1$ collapses $R_v$ regardless of
true convergence, and cross-cohort agreement on a single period's
level need not imply that the projection has settled. The dual
criterion closes both inertia-leakage paths.

## Why two conditions

A **denominator effect** disables any single-criterion diagnostic.

In long-duration health insurance, cumulative LR = cumulative loss /
cumulative risk premium. As dev grows, the denominator grows
alongside the numerator, so a new calendar diagonal's contribution to
the overall ratio shrinks automatically — *regardless of whether the
underlying experience has actually changed*. This is the *inertia*
effect.

What each criterion guards against:

| Scenario | $R_v$ | $\hat{D}_v$ | Result |
|---|---|---|---|
| True convergence (model + experience both stable) | small | small | **PASS** ✓ |
| Chain-ladder $\hat{f}_k \to 1$ drift (inertia) | small (spurious) | large | FAIL — dispersion catches it |
| Coincidental cohort agreement at one period | large | small (snapshot) | FAIL — projection revision catches it |

- $R_v$ alone is fooled by chain ladder's mechanical drift
  ($\hat{f}_k \to 1$): the cumulative product barely moves, the
  projection barely moves, and $R_v$ collapses to zero — a false
  convergence.
- $\hat{D}_v$ uses **incremental** LR, so it inherits no cumulative
  denominator → immune to the inertia effect.
- Requiring both criteria simultaneously closes the principal
  inertia-leakage paths. That is the design intent of the dual
  criterion.

## Notation

| Symbol | Meaning |
|--------|---------|
| $i$ | cohort index (UY) |
| $v$ | valuation index — the calendar diagonal; "$v$ diagonals observed" |
| $V$ | maximum observed valuation (max dev in the triangle) |
| $k^*$ | maturity point (from `detect_maturity()`); lower bound on candidate $v$ |
| $k^{**}$ | convergence point — the value `detect_convergence()` returns |
| $\hat{LR}^{\mathrm{proj}}_v$ | projected ultimate LR using data through valuation $v$ |
| $R_v$ | revision: $\lvert\hat{LR}^{\mathrm{proj}}_v - \hat{LR}^{\mathrm{proj}}_{v-1}\rvert$ |
| $\hat{SE}^{\mathrm{param}}_v$ | parameter-uncertainty SE of $\hat{LR}^{\mathrm{proj}}_v$ (Mack-style) |
| $\hat{lr}_{i,v}$ | incremental loss ratio of cohort $i$ at dev $v$ |
| $\hat{D}_v$ | robust scale-invariant dispersion of $\hat{lr}_{i,v}$ across cohorts |
| $c$ | multiplier on $\hat{SE}^{\mathrm{param}}_v$ for the revision gate (default `0.5`) |
| $\tau$ | upper bound on $\hat{D}_v$ for the dispersion gate (default `0.15`) |
| $M$ | required run length of consecutive passing valuations (default `3L`) |

The constant $1.4826 \approx 1 / \Phi^{-1}(0.75)$ inside $\hat{D}_v$ is
the standard MAD$\to\sigma$ correction: with this scaling
$\mathrm{MAD}_i$ becomes a consistent estimator of the cross-cohort
standard deviation under normality, so $\hat{D}_v$ reads as a robust,
outlier-resistant coefficient of variation of incremental LR.

## Basic usage

```{r, eval = FALSE}
library(lossratio)
data(experience)
exp <- as_experience(experience)
tri <- build_triangle(exp[cv_nm == "SUR"], cv_nm)

res <- detect_convergence(tri)
print(res)
```

Mock output:

```
#> <Convergence>
#> k_conv       : NA
#> k_star       : 9
#> V (max dev)  : 30
#> criterion    : R_v < 0.5 * SE_param_v  AND  D_v < 0.15  (run M = 3)
#> fit_fn       : fit_lr
#> v candidates : 19 ( 0  pass both clauses)
```

The returned `Convergence` object reports:

- `k_conv` — the detected $k^{**}$, or `NA` if no run of $M$
  consecutive passing valuations is found.
- `k_star` — the maturity point used as the lower bound (computed
  internally via `detect_maturity()` on a clr-based ATA, or supplied
  by the caller).
- `V` — the maximum observable dev in the triangle.
- `v`, `R_v`, `SE_param_v`, `D_v`, `pass_v` — per-valuation diagnostic
  sequences indexed by $v$.
- `c`, `tau`, `M`, `holdout_max`, `min_n_cohorts` — settings used.
- attributes `group_var`, `value_var`, `fit_fn_name`, `dev_var`.

`summary(res)` returns a `data.table` with one row per candidate
valuation and an extra `R_over_SE = R_v / SE_param_v` column for
inspection:

```{r, eval = FALSE}
head(summary(res), 6)
```

```
#>        v    R_v   SE_param_v  R_over_SE   D_v     pass
#> 1:     9     NA           NA         NA  0.90   FALSE
#> 2:    10     NA           NA         NA  0.76   FALSE
#> 3:    11     NA           NA         NA  0.56   FALSE
#> 4:    12     NA           NA         NA  0.58   FALSE
#> 5:    13     NA           NA         NA  0.81   FALSE
#> 6:    14     NA           NA         NA  0.43   FALSE
```

## How it works: multiple holdout refits

`detect_convergence()` refits the model at each candidate valuation
and tracks how the projection changes.

Example: with $V = 30$, $k^* = 18$, and `holdout_max = 6` —
candidates are $v \in \{24, 25, \dots, 30\}$ (7 in total).

| $v$ | holdout depth ($V - v$) | $R_v$ available? |
|---|---|---|
| 30 | 0 | ✓ |
| 28 | 2 | ✓ |
| 24 (cutoff) | 6 | ✓ |
| 22 | 8 | `NA` |
| 18 | 12 | `NA` |

One refit per candidate $v$ (7 total) plus $R_v$ computed between
adjacent $v$. The cutoff `holdout_max` defaults to
`floor((V - k_star) / 2)`; once holdout depth exceeds it the refit
data becomes too thin and $R_v$, $SE_param_v$ are masked to `NA`.

Increase `holdout_max` to diagnose deeper into the past — at the
cost of less data behind each refit, hence lower confidence.

## Plot

```{r, eval = FALSE}
plot(res)
```

The diagnostic is two stacked panels: the upper panel shows
$R_v / \hat{SE}^{\mathrm{param}}_v$ against $v$ with a horizontal
guide at $c$; the lower panel shows $\hat{D}_v$ against $v$ with a
horizontal guide at $\tau$. A vertical dotted line marks $k^*$, and a
vertical solid line marks $k^{**}$ when one is detected.
A point falling **below both threshold lines** passes the joint
criterion.

This view is also a quick way to see *which clause is binding*. If the
top panel hugs the threshold but the bottom is far above, the issue is
cross-cohort heterogeneity; if the bottom is fine but the top is high,
the model is still revising.

## Threshold tuning

The defaults are deliberately conservative:

| Argument | Default | Meaning |
|----------|---------|---------|
| `c`      | `0.5`   | Revision must be smaller than half the parameter SE. |
| `tau`    | `0.15`  | Cross-cohort dispersion must be below 15% of the median lr. |
| `M`      | `3L`    | Both clauses must hold for at least 3 consecutive valuations. |
| `min_n_cohorts` | `5L` | Below this cohort count, $\hat{D}_v$ is `NA` (insufficient sample). |

Tighter thresholds yield later (or no) $k^{**}$; sweep a
range to inspect sensitivity:

```{r, eval = FALSE}
sapply(
  c(0.25, 0.5, 0.75, 1.0),
  function(cc) detect_convergence(tri, c = cc)$k_conv
)
```

Values of $\hat{D}_v$ below $\tau \approx 0.05$ are difficult to
attain in real portfolios because of single-period claim noise; values
above $0.20$ usually indicate genuine cohort heterogeneity that
warrants `detect_regime()` before further modelling.

## Relation to `k^*` and `detect_regime()`

The three diagnostics answer different questions and operate on
different axes:

| Tool | Question | Result | Axis |
|------|----------|--------|------|
| `detect_regime()` | Are cohorts homogeneous? | cohort groups | underwriting period |
| `detect_maturity()` ($k^*$) | When are link factors reproducible? | a dev value | development period |
| `detect_convergence()` ($k^{**}$) | When does the LR estimate stop revising? | a dev value | development period |

A defensible workflow is:

1. Run `detect_regime()`. If multiple regimes exist, fit each
   group separately.
2. For each homogeneous group, compute $k^*$ via
   `detect_maturity()`.
3. Run `detect_convergence()` to obtain $k^{**} \ge k^*$. The reported
   $\hat{lr}$ at the convergence point is the projected ultimate loss
   ratio averaged over $k \ge k^{**}$ (or read directly from
   `fit_lr()$summary`).

The sequence separates *cohort homogeneity*, *link reproducibility*,
and *level convergence* — three properties that coincide in P&C
run-off but must be verified independently in long-duration health
insurance.

## Limitations

`detect_convergence()` is a thin layer over repeated `backtest()` calls
and inherits their constraints:

- **Identifiability**: $k^{**}$ can be declared only when $V \ge k^* + M$;
  short observation windows return `NA`.
- **Model conditioning**: $\hat{LR}^{\mathrm{proj}}_v$ is computed
  by `fit_fn` (default `fit_lr`). Different fitters yield different
  $k^{**}$. Reporting under multiple `fit_fn` is recommended for
  robustness.
- **Portfolio aggregation**: $R_v$ and $\hat{SE}^{\mathrm{param}}_v$
  are exposure-weighted across cohorts assuming inter-cohort
  independence. Calendar-year shocks (regulatory, healthcare cost
  trend) violate this assumption; both clauses can move together for
  non-cohort reasons.
- **Multi-group triangles**: the helper currently collapses
  $\hat{D}_v$ across groups by median; running each group
  separately is recommended when groups behave differently.

## See also

- `?detect_convergence`, `?detect_maturity`, `?backtest`,
  `?detect_regime`.
- Vignette `regime-detection` — cohort homogeneity diagnostics that
  feed step 1 of the workflow above.
