---
title: "Convergence: detecting when projected loss ratio stops revising"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence: detecting when projected loss ratio stops revising}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

## Motivation

`detect_maturity()` answers *"from which development period are link
factors $f_k$ reproducible across cohorts?"*. That is necessary for
chain-ladder projection but not sufficient for declaring a portfolio's
projected loss ratio converged. In long-duration health insurance both
$f_k \to 1$ and $g_k \to 0$ arise mechanically as cumulative
denominators grow — independent of whether the underlying experience
has actually settled. A criterion built on those quantities passes
automatically with $k$, not because of true convergence (the *inertia*
failure mode).

`detect_convergence()` detects the **convergence point** $k^{**}$ —
the first dev $k \ge k^*$ at which the projected portfolio loss ratio
is observed to be stable, in a sense the user picks via `method =`. It
is the natural counterpart to $k^*$:

- $k^*$ (`detect_maturity()`) marks where link factors $f_k$ become
  reproducible.
- $k^{**}$ marks where the projection itself stops moving with new
  data.

Long-duration health portfolios may cross $k^*$ early yet remain far
from $k^{**}$.

## The four stability methods

Convergence cannot be measured asymptotically from finite data, only
*observed up to* the maximum available dev `dev_max` ($K_{\max}$). The
detector runs a rolling `backtest()` over a sequence of candidate dev
points `dev_cand` $\in [k^*, K_{\max}-2]$, builds the projected
ultimate LR path `lr` at each one, then evaluates four stability
metrics on that path. The `method =` argument selects which metric
defines `conv_k`; all four are returned for inspection regardless of
the choice.

| Method | Metric | Captures | Misses |
|---|---|---|---|
| `"window"` | range of `lr` over the next `window` valuations | local stability (no zig-zag) | slow monotone drift below `max_drift` per window |
| `"tail"` (default) | range of `lr` over `[k, K_{\max}]` | global stability up to `dev_max`; catches monotone drift | needs $\ge 2$ tail points; first-pass is conservative (later than `"window"`) |
| `"slope"` | $\|\hat\beta_k\|$, OLS slope of `lr ~ k` on `[k, K_{\max}]` | systematic trend (signed direction) | oscillation that averages out to zero slope |
| `"all"` | intersection of `"window"` + `"tail"` + `"slope"` | every shape above | (strictest) |

Across all methods a cross-cohort dispersion clause `dispersion[i] <
max_dispersion` is required in addition — incremental loss ratios at
that dev must agree across cohorts (computed via the same robust
$\hat{D}_v$ metric used in earlier package versions).

**Why not the paper's SE-normalised criterion?** Earlier versions of
this detector implemented $R_k < c \cdot \hat{SE}^{\mathrm{param}}_k$
from the original methodology paper (Section 11; paper uses $v$
instead of $k$, same axis). On large portfolios that form is
structurally broken: $\hat{SE}^{\mathrm{param}}$ shrinks as
$1/\sqrt{n}$ while $R_k$ has a numerical noise floor (~$10^{-3}$
LR units), so the ratio diverges and the criterion never fires —
even on synthetic data that is *visibly stable*. The drift-based
methods replace SE-normalisation with a data-size-independent
threshold.

## Notation

Standard chain-ladder convention: $i$ indexes cohort (origin period),
$k$ indexes development period. `detect_maturity()` returns $k^*$,
`detect_convergence()` returns $k^{**}$ — both live on the same $k$
axis.

| Code | Math | Meaning |
|---|---|---|
| `dev_max` | $K_{\max}$ | Maximum observable dev (a scalar) |
| `dev_cand` | $k \in [k^*, K_{\max}-2]$ | Integer vector of candidate dev points |
| `lr[i]` | $LR_k$ | Portfolio LR projection at dev = `dev_cand[i]` |
| `revision[i]` | $R_k = \|LR_k - LR_{k-1}\|$ | Adjacent-step revision (diagnostic only) |
| `drift_window[i]` | $\max - \min$ over $[k, k+W-1]$ | Local window range |
| `drift_tail[i]`   | $\max - \min$ over $[k, K_{\max}]$ | Tail range (global stability) |
| `slope[i]` | $\hat\beta_k$ | OLS slope on $[k, K_{\max}]$ |
| `dispersion[i]` | $\hat{D}_k$ | Robust cross-cohort spread of incremental LR |
| `mat_k` | $k^*$ | Maturity point (lower bound on candidates) |
| `conv_k` | $k^{**}$ | The detected convergence point |

The constant $1.4826 \approx 1 / \Phi^{-1}(0.75)$ inside
`dispersion` is the standard MAD$\to\sigma$ correction. With this
scaling, $\hat{D}_k$ reads as a robust, outlier-resistant coefficient
of variation of incremental LR.

## Basic usage

```{r, eval = FALSE}
library(lossratio)
data(experience)
tri <- as_triangle(
  experience[coverage == "surgery"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_prem"
)

res <- detect_convergence(tri)
print(res)
```

Mock output (no convergence detected on this run-off):

```
#> <Convergence>
#>   method     : tail
#>   conv_k     : NA
#>   mat_k      : 4
#>   dev_max    : 30
#>   candidates : 25
#>   passes :
#>     window :  0/25 (drift_window < 0.01  & dispersion < 0.15)
#>     tail   :  0/25 (drift_tail   < 0.01  & dispersion < 0.15)  <- method
#>     slope  :  0/25 (|slope|      < 0.001 & dispersion < 0.15)
#>     all    :  0/25 (window AND tail AND slope)
```

The summary `data.table` has one row per candidate dev with every
metric and per-method pass flag:

```{r, eval = FALSE}
head(summary(res), 6)
```

```
#>      dev    lr   revision  drift_window  drift_tail   slope  dispersion
#>    <int> <num>      <num>         <num>       <num>   <num>       <num>
#> 1:     4 0.62          NA          0.07        0.07   0.001        0.47
#> 2:     5 0.63        0.01          0.06        0.06   0.001        0.47
#> ...
#>    pass_window  pass_tail  pass_slope  pass
#>          <lgl>      <lgl>       <lgl> <lgl>
#> 1:       FALSE      FALSE       FALSE FALSE
#> 2:       FALSE      FALSE       FALSE FALSE
```

The `Convergence` object also reports the threshold parameters
(`max_drift`, `max_slope`, `max_dispersion`, `window`,
`holdout_max`, `min_n_cohorts`) and metadata attributes (`groups`,
`target`, `dispatcher`).

## Reserving caveat

**The detected `conv_k` is stability *observed up to* `dev_max`, not
an asymptotic guarantee.** Future development past `dev_max` is
unknown. Treat `conv_k` as a diagnostic for "from here on, what we
observe is stable", not as proof that the projection will not drift
later.

For reserving applications:

- Prefer `method = "tail"` (default) or `"all"` over `"window"`.
  `"window"` will declare convergence prematurely on slow monotone
  drifts that fit under `max_drift` per window — exactly the
  silent-revision pattern that hurts reserves.
- Always weigh the **evidence span** `dev_max - conv_k`. A `conv_k`
  near `dev_max` (say, span $< 5$) was confirmed by very few tail
  points; the same data might unconverge with one more diagonal.
- Read the projected ultimate LR and its standard error from
  `fit_lr()$summary` directly. `detect_convergence()` is a diagnostic,
  not an estimator; the reserve point estimate and uncertainty come
  from the fit object.

## How it works: rolling holdout refits

For each `dev_cand[i]`, the function backtests the LR projection with
`holdout = dev_max - dev_cand[i]` and extracts the portfolio LR. Calls
are cached per holdout depth so we don't redo work between adjacent
candidates.

Example: with `dev_max = 30`, `mat_k = 4`, `holdout_max = 13` —
candidates are $k \in \{4, 5, \dots, 28\}$ (25 in total), but only
$k$ with `holdout = dev_max - k <= holdout_max` get a finite `lr[i]`.
The rest are masked to `NA`.

`holdout_max` defaults to `max(window, floor((dev_max - mat_k) / 2))`.
Raising it widens the diagnostic window — at the cost of less data
behind each refit and hence noisier `lr` near the lower edge.

## Plot

```{r, eval = FALSE}
plot(res)
```

Five stacked panels share the dev axis:

1. **`lr`** — the LR trajectory the metrics are computed from.
2. **`drift_window`** — the local-window metric, with horizontal
   guide at `max_drift`.
3. **`drift_tail`** — the tail metric, same guide.
4. **`|slope|`** — guide at `max_slope`.
5. **`dispersion`** — guide at `max_dispersion`.

Vertical guides mark `mat_k` (dashed) and `conv_k` (solid, only when
non-`NA`). For each metric panel, points below the red dashed line
pass that clause. The subtitle records the active `method`. This is
also a quick way to see *which clause is binding*: any panel that
hovers above its threshold blocks convergence under that method.

## Threshold tuning

| Argument | Default | Meaning |
|----------|---------|---------|
| `method`         | `"tail"` | Which stability metric defines `conv_k`. |
| `max_drift`      | `0.01`   | Drift cap on `drift_window` / `drift_tail`, in LR units. Raise (e.g. `0.02`–`0.05`) for noisier or longer-tail books. |
| `max_slope`      | `1e-3`   | $\|\hat\beta_k\|$ cap (LR per dev). |
| `max_dispersion` | `0.15`   | Cross-cohort dispersion cap. |
| `window`         | `5L`     | Drift window length $W$ — number of consecutive valuations the `"window"` method scans when computing `drift_window`. Does *not* affect `"tail"` or `"slope"`. |
| `min_n_cohorts`  | `5L`     | Below this cohort count, `dispersion` is `NA`. |

Sweep `max_drift` to inspect sensitivity:

```{r, eval = FALSE}
sapply(
  c(0.005, 0.01, 0.02, 0.05),
  function(d) detect_convergence(tri, method = "tail", max_drift = d)$conv_k
)
```

`max_dispersion` below $\approx 0.05$ is difficult to attain in real
portfolios because of single-period claim noise; values above $0.20$
usually indicate genuine cohort heterogeneity that warrants
`detect_regime()` before further modelling.

## Relation to $k^*$ and `detect_regime()`

The three diagnostics answer different questions on different axes:

| Tool | Question | Result | Axis |
|------|----------|--------|------|
| `detect_regime()` | Are cohorts homogeneous? | cohort groups | underwriting period |
| `detect_maturity()` ($k^*$) | When are link factors reproducible? | a dev value | development period |
| `detect_convergence()` ($k^{**}$) | When does the LR estimate stop revising? | a dev value | development period |

A defensible workflow:

1. Run `detect_regime()`. If multiple regimes exist, fit each
   group separately (or use `regime_at()` / `regime_spec()` in the
   fit / backtest call).
2. For each homogeneous group, compute $k^*$ via `detect_maturity()`.
3. Run `detect_convergence()` to obtain $k^{**} \ge k^*$. Read the
   projected ultimate loss ratio from `fit_lr()$summary` and apply
   the reserving caveat above.

The sequence separates *cohort homogeneity*, *link reproducibility*,
and *level convergence* — three properties that coincide in P&C
run-off but must be verified independently in long-duration health
insurance.

## Limitations

`detect_convergence()` is a thin layer over repeated `backtest()`
calls and inherits their constraints:

- **Identifiability**: `conv_k` can be declared only when `dev_max -
  mat_k >= window` (or 2 for tail / slope). Short observation windows
  return `NA` for every method.
- **Model conditioning**: the projected LR is computed by `fit_lr()`,
  which internally composes `fit_loss()` (default `method = "sa"`
  — stage-adaptive) and `fit_premium()`. Choices made inside that
  composition (loss method, regime filter, maturity argument) feed
  through to `conv_k`; review the underlying `fit_lr` settings when
  interpreting the result. Pass `loss_method =`, `loss_regime =`,
  etc. through `...` to override.
- **Portfolio aggregation**: the portfolio LR aggregates per-group
  ultimates via exposure weighting (`sum(loss_ult) /
  sum(premium_ult)`). Calendar-year shocks (regulatory changes, cost
  trend) can move every group together; the drift metrics will not
  separate that from genuine convergence.
- **Multi-group triangles**: `dispersion` is currently collapsed
  across groups by median. When groups behave differently, running
  each group separately is recommended.
- **Asymptotic vs observed**: as flagged in the reserving caveat,
  every method only measures stability over the available history.
  Confirmed silent drift past `dev_max` is impossible to detect from
  these tools alone.

## See also

- `?detect_convergence`, `?detect_maturity`, `?backtest`,
  `?detect_regime`, `?fit_lr`.
- `vignette("regime-detection")` — cohort homogeneity diagnostics that
  feed step 1 of the workflow above.
- `vignette("backtest")` — the underlying rolling holdout machinery
  that `detect_convergence()` is built on.
