---
title: "Get started: lossratio quick tour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started: lossratio quick tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

This vignette walks the full `lossratio` pipeline on the bundled
synthetic experience data, from raw long-format rows to a fitted
loss-ratio projection.

## Input shape

`lossratio` consumes long-format experience data — one row per
(cohort × dev × group) cell. The bundled dataset
`experience` is a 2,664-row table with calendar / underwriting
period columns at multiple granularities, a `coverage` group
column, and per-period amounts (`loss_incr`, `premium_incr`).

```{r}
library(lossratio)

data(experience)
str(experience)
```

## Step 1 — Validate and coerce

```{r}
exp <- as_experience(experience)
class(exp)
```

`as_experience()` checks required columns (`cy_m`, `uy_m`, `loss_incr`,
`premium_incr`), coerces date columns, and tags the class. Use
`check_experience(df)` for a non-mutating check.

## Step 2 — Build the cohort × dev structure

```{r}
tri <- build_triangle(exp[coverage == "SUR"], group_var = coverage)
class(tri)
names(tri)
```

`build_triangle()`:

- aggregates demographic dimensions away (here: `age_band`, `gender`),
- standardises raw input names to the per-period slots `loss_incr`
  and `premium_incr`,
- adds cumulative companions (`loss`, `premium`) — cumulative is the
  unmarked default; per-period carries the `_incr` suffix,
- adds derived metrics (`margin` / `margin_incr`, `lr` / `lr_incr`,
  proportions),
- standardises the cohort / development columns to the names `cohort`
  and `dev`,
- preserves original column names as attributes (`cohort_var`,
  `dev_var`, `loss_var`, `premium_var`) for downstream plot labels.

## Step 3 — Diagnostics

```{r}
plot(tri)              # cohort trajectories of lr
plot_triangle(tri)     # heatmap of lr cells
summary(tri)           # group-wise statistics by dev
```

## Step 4 — Link-level diagnostics

Two complementary views of the per-link factor that drives development.
Both are *factor-level* — they return per-link estimates only; they do
not produce projections (`$full`).

```{r}
# Age-to-age factors — multiplicative, input to chain ladder
fit_ata(tri, loss_var = "loss")

# Exposure-driven intensities — additive, input to ED
fit_intensity(tri, loss_var = "loss", premium_var = "premium")
```

`fit_ata()` returns selected ATA factors $f_k$ per development link.
`fit_intensity()` returns intensity factors $g_k = \Delta C^L_k / C^P_k$.
Both feed the projection-level estimators below and are used to detect
the maturity point.

## Step 5 — Projection

The projection-level estimators come as a pair of siblings,
`fit_cl()` and `fit_ed()`, plus `fit_lr()` as the unified interface
that combines them.

```{r}
# Chain ladder (multiplicative projection)
cl <- fit_cl(tri, loss_var = "loss", method = "mack")
plot(cl, type = "projection")
summary(cl)

# Exposure-driven (additive projection)
ed <- fit_ed(tri, loss_var = "loss", premium_var = "premium")
summary(ed)
```

`fit_lr()` is the unified loss-ratio interface. The default
`method = "sa"` (stage-adaptive) uses exposure-driven before the
maturity point and chain ladder after, switching at the per-group
maturity detected from the ATA factors. `method = "ed"` or `"cl"`
applies a single stage throughout.

```{r}
lr <- fit_lr(tri, method = "sa")
plot(lr, type = "lr")
summary(lr)
```

## Step 6 — Structural change diagnostics

`detect_regime()` checks whether recent cohorts behave
differently from earlier ones. Useful as a preprocessing step before
loss-ratio fitting on a homogeneous subset:

```{r}
sub <- build_triangle(exp[coverage == "SUR"], group_var = coverage)
detect_regime(sub, K = 12, method = "e_divisive")
```

See `vignette("regime")` for the dedicated walkthrough.

## Where to next

- `vignette("aggregation-frameworks")` — when to use `build_triangle`
  vs `build_calendar` vs `build_total`.
- `vignette("projection")` — choosing among `"sa"` / `"ed"` /
  `"cl"` in `fit_lr()`.
- `vignette("chain-ladder-reserving")` — `fit_cl()` deep dive (Mack variance,
  tail factor).
- `vignette("triangle-link-and-maturity")` — `summary()`,
  `detect_maturity()`, and triangle-style visualisations.
- `vignette("regime")` — `detect_regime()` for
  structural-change diagnosis across cohorts.
- `vignette("backtest")` — calendar-diagonal hold-out validation with
  `fit_cl` or `fit_lr`.
