---
title: "Getting started with lossratio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lossratio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
options(lossratio.font = "")
library(lossratio)
```

This vignette walks the core functions of the lossratio package end to
end, in the order **prepare the data -> fit -> diagnose -> validate ->
visualise**. At each step it explains *what* is being called, *why* it
behaves the way it does, and *how* to read the result. Reading this one
document is enough to grasp the full workflow.

## The problem this package solves

In long-term health insurance a policy, once underwritten, stays in
force for decades. The central task is therefore to look ahead and see
how a group of contracts written at a given point in time (a cohort)
accumulates its loss ratio as the years pass.

lossratio frames this problem on a **cohort x development-period
(cohort x dev)** grid.

- **cohort** -- the time a contract was underwritten (e.g. January 2023
  business).
- **development period (dev)** -- the time elapsed since underwriting.

Each cell holds a cumulative loss (`loss`) and a cumulative premium
(`premium`), and the loss ratio is defined as `ratio = loss / premium`.
Estimating the cells of future development periods that have not yet
been observed is the act of projection, and the various `fit_*`
functions in this package perform that estimate in different ways.

## The data: `experience`

The package ships with a synthetic experience dataset. The term
experience is the domain standard, referring to the cell-level records
of an actuarial experience study.

```{r data}
data(experience)
str(experience)
head(experience)
```

The key columns are:

- `coverage` -- the line of business (group). Four products: surgery,
  ci, cancer, death.
- `uy_m` -- underwriting month. The source column for the cohort axis.
- `cy_m` -- calendar month.
- `dev_m` -- development month.
- `incr_loss`, `incr_premium` -- **incremental (per-period)** loss and
  premium.

Note that the input is incremental: raw experience is usually
accumulated in the form "how much was incurred / earned this month".
The cumulative values are constructed inside the package.

## Building a Triangle -- `as_triangle()`

`as_triangle()` turns raw experience into a standard **Triangle**
object. It performs validation, coordinate standardisation, and
aggregation in one call.

```{r triangle}
tri <- as_triangle(
  experience,
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_premium"
)
tri
```

It takes the original column names (`uy_m`, `incr_loss`, ...) and
renames them to standard names (`cohort`, `dev`, `loss`, `premium`,
...), preserving the originals as attributes. The output columns are
**cumulative by default**, with incremental columns carrying the
`incr_` prefix -- `loss`/`incr_loss`, `premium`/`incr_premium`,
`ratio`/`incr_ratio`.

If you already hold a cumulative triangle, pass `cell_type =
"cumulative"`. `as_triangle()` then differences within each cohort to
recover the increments automatically.

```{r triangle-cumulative, eval = FALSE}
# when loss / premium are already cumulative
as_triangle(df_cumulative,
            groups = "coverage", cohort = "uy_m", calendar = "cy_m",
            loss = "loss", premium = "premium",
            cell_type = "cumulative")
```

The rest of the walkthrough uses a single product so the output stays
compact.

```{r triangle-one}
tri1 <- as_triangle(
  experience[coverage == "surgery"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_premium"
)
```

The triangle itself can be viewed as a heatmap.

```{r triangle-plot}
plot_triangle(tri1, metric = "ratio")
```

Only the upper-left region (the observed area) is filled; the empty
lower-right cells are the future cells still to be projected.

## Two views of loss development

To project future cells, the way "loss grows over development time"
must be expressed as a model. lossratio uses two basic views.

- **Multiplicative -- chain ladder (CL)**. The cumulative loss at the
  next development period is a multiple of the current one: `C_{k+1} =
  f_k * C_k`. Here `f_k` is the age-to-age factor, estimated as the
  ratio of cumulative loss between two adjacent development periods.
- **Additive -- exposure-driven (ED)**. The loss *increment* of the
  next period is made proportional to premium: `Δloss = g_k * premium`.
  `g_k` is the loss intensity per unit of premium.

The two are complementary. Early in development the cumulative loss is
small, so the age-to-age factor swings widely; anchoring on premium
makes the exposure-driven view more stable. As development matures the
age-to-age factor settles, and chain ladder preserves the cohort's own
observed level well.

## `fit_cl()` -- chain ladder

`fit_cl()` performs a Mack (1993) chain ladder fit on the cumulative
`loss`.

```{r fit-cl}
cl <- fit_cl(tri1)
summary(cl)
```

`summary()` reports the ultimate loss per cohort with its standard
error (`loss_total_se`) and coefficient of variation (`loss_total_cv`).
The standard error follows Mack's variance decomposition, computed as
the sum of *process error* and *parameter error*.

The `$full` slot holds the cell-level result over both the observed and
projected regions.

```{r fit-cl-full}
head(cl$full[, c("cohort", "dev", "loss_obs", "loss_proj",
                 "loss_total_se", "is_observed")])
```

The projection curves connect each cohort's observed values (solid)
with its projection (dashed).

```{r fit-cl-plot, fig.height = 8}
plot(cl)
```

## `fit_ed()` -- exposure-driven

`fit_ed()` fits the same triangle with the additive exposure-driven
model.

```{r fit-ed}
ed <- fit_ed(tri1)
summary(ed)
```

The point projection (`loss_proj`) is nearly identical between chain
ladder and exposure-driven -- both explain the same data. The
difference lies in *how the variance accumulates*. Chain ladder
amplifies variance through the square of the multiplier `f`, while the
exposure-driven model adds premium-proportional increments. As a
result the exposure-driven standard error is calmer in the early
development periods.

## `fit_sa()` -- stage-adaptive

`fit_sa()` blends the two views in a stage-adaptive (SA) way: the
development periods before the maturity point are fit with the
exposure-driven model, and the periods after it with chain ladder.

```{r fit-sa}
sa <- fit_sa(tri1)
summary(sa)
```

The maturity point is the development period where the age-to-age
factor has settled sufficiently, detected automatically inside the fit
(two-pass). The intuition is simple: the early region, where the
age-to-age factor still swings, anchors on premium (ED); the settled
region thereafter follows the cohort's own development (CL). Stage
adaptation is the package's default workhorse for long-term loss-ratio
projection.

```{r fit-sa-plot, fig.height = 8}
plot(sa)
```

## `fit_bf()` -- Bornhuetter-Ferguson

`fit_bf()` brings in an **external prior loss ratio**. The portion of
loss that has not yet emerged is filled not by the data but by the
prior.

\[
\hat L^{ult}_i = L^{obs}_i + (1 - q_i)\cdot \mathrm{ELR}_i \cdot E^{ult}_i
\]

`q_i` is the emergence ratio of cohort i (`L^{obs} / L^{ult, CL}`). In
other words Bornhuetter-Ferguson is a credibility blend that fills the
*already-emerged part with data* and the *not-yet-emerged part with the
prior*. The `prior` argument is required.

```{r fit-bf}
bf <- fit_bf(tri1, prior = 0.70)
summary(bf)
```

The prior can vary by cohort (a data frame) or be a distributional
prior (with an `elr_se` column). The younger a cohort -- the less data
it has -- the more its result is pulled toward the prior; that is
precisely the reason to use Bornhuetter-Ferguson.

## `fit_cc()` -- Cape Cod

`fit_cc()` has the same structure as Bornhuetter-Ferguson, except the
prior loss ratio is not supplied externally -- it is **estimated by
pooling the data**.

\[
\widehat{\mathrm{ELR}} =
  \frac{\sum_i L^{obs}_i}{\sum_i E^{ult}_i \cdot q_i}
\]

```{r fit-cc}
cc <- fit_cc(tri1)
summary(cc)
```

Apart from the absence of a `prior` argument, the call and the output
match Bornhuetter-Ferguson. It is the method to use when there is no
external basis for a prior and you let the portfolio speak its own ELR.

## `fit_loss()` / `fit_premium()` -- the unified dispatchers

Rather than calling a different function for each method, `fit_loss()`
lets you switch the method with a single `method` argument.

```{r fit-loss}
fl <- fit_loss(tri1, method = "sa")
class(fl)
```

`method` is one of `"ed"` (default), `"cl"`, `"sa"`, `"bf"`, `"cc"`.
For `"bf"` / `"cc"` extra arguments such as `prior` are forwarded
through `...`. `fit_premium()` is the companion function that fits the
premium side separately.

```{r fit-premium}
fe <- fit_premium(tri1)
summary(fe)
```

## `fit_ratio()` -- composing the loss ratio

`fit_ratio()` is the composition layer: it fits the loss side and the
premium side separately, then combines them to produce the **loss
ratio**.

```{r fit-ratio}
lr <- fit_ratio(tri1, method = "sa")
summary(lr)
```

`$full` holds `loss_proj` / `premium_proj` / `ratio_proj` together with
the loss-ratio standard error (`ratio_se`) and confidence interval
(`ratio_ci_lo` / `ratio_ci_hi`).

```{r fit-ratio-full}
head(lr$full[, c("cohort", "dev", "loss_proj", "premium_proj",
                 "ratio_proj", "ratio_se")])
```

The loss-ratio standard error is governed by `se_method`: `"fixed"`
(default) treats premium as a known value, while `"delta"` carries the
uncertainty of premium and the loss-premium correlation through the
delta method.

```{r fit-ratio-plot, fig.height = 8}
plot(lr)
```

## Calendar and Total -- the other aggregation frameworks

The same long-format experience can be aggregated three ways depending
on the question. `as_triangle()` keeps both axes (cohort x dev) for
projection; two siblings collapse one or both axes.

`as_calendar()` collapses cohorts onto the diagonal, giving one row per
calendar period -- the diagonal sum of the triangle. Use it for
calendar-time trend and diagonal-effect analysis.

```{r as-calendar}
cal <- as_calendar(tri1)
head(cal)
plot(cal)              # x axis: calendar (Date)
```

The `t` column is a sequential index within group (1, 2, 3, ...), a
time-series convention. It is *not* a development period.

`as_total()` collapses both axes to one value per group, useful for
portfolio-level comparison. The `period_from` / `period_to` arguments
bound a fixed window so groups stay comparable.

```{r as-total}
tot <- as_total(
  as_triangle(
    experience[uy_m >= as.Date("2023-04-01") &
               uy_m <= as.Date("2024-03-01")],
    groups = "coverage", cohort = "uy_m", dev = "dev_m",
    loss = "incr_loss", premium = "incr_premium"
  )
)
head(tot)
```

## The Link object -- factor-level diagnostics

Between a Triangle and a fit sits the **Link** object, the per-link
factor table built by `as_link()`. It is the intermediate object that
feeds both chain ladder and the exposure-driven model.

```{r as-link}
# single-variable mode -> observed age-to-age factors
ata <- as_link(tri1, loss = "loss")
summary(ata, model = "ata")

# dual-variable mode -> exposure-driven intensities g_k
ed_link <- as_link(tri1, loss = "loss", exposure = "premium")
summary(ed_link, model = "ed")
```

In single-variable mode (`loss` only) the Link carries the observed
age-to-age factors. Supplying `exposure` adds the exposure-driven
intensity `intensity = loss_delta / premium_from`. The `summary()`
method computes the per-link statistics -- mean / median / weighted
factor, coefficient of variation, WLS standard error -- that drive
maturity detection. The factor-level helpers `fit_ata()` and
`fit_intensity()` wrap this same machinery and return per-link
estimates only (no `$full` projection).

## Diagnostics -- maturity, convergence, regime

There are diagnostic functions for inspecting the triangle's structure
before and after fitting.

**Maturity point** -- the development period where the age-to-age
factor settles. It is also the stage-switch point of the
stage-adaptive method.

```{r detect-maturity}
detect_maturity(tri1)
```

**Convergence point** -- the development period beyond which the
projected loss ratio no longer moves meaningfully. As the denominator
of the cumulative loss ratio grows it dampens the signal -- an inertia
effect -- which a dual criterion is used to filter through.

```{r detect-convergence}
detect_convergence(tri1)
```

**Regime** -- a structural change between underwriting cohorts. It
detects whether a premium, policy-wording, or underwriting-guideline
change altered the loss behaviour of cohorts after a certain point.

```{r detect-regime}
reg <- detect_regime(tri)
reg$changes
```

The surgery product in `experience` has a synthetic regime change
planted during 2024, so the result above picks up that change point.

## `backtest()` -- validation

`backtest()` hides the latest few calendar diagonals, refits the model,
and compares the withheld actuals against the projection to measure the
model's accuracy.

```{r backtest}
bt <- backtest(tri1, holdout = 6L, target = "ratio", loss_method = "sa")
summary(bt)
```

`target` is one of `"ratio"` / `"loss"` / `"premium"`, calling
`fit_ratio` / `fit_loss` / `fit_premium` internally. The resulting A/E
error tables show which development periods and which diagonals the
model missed.

```{r backtest-plot}
plot(bt)
```

## Visualisation -- `plot()` / `plot_triangle()`

Every major object supports two families of plots.

- `plot(x)` -- trajectory / diagnostic charts (per-cohort development
  curves, regime PCA, and so on).
- `plot_triangle(x)` -- triangular heatmaps (cell values or cell
  status).

```{r plot-examples, fig.height = 8}
plot_triangle(sa, label_style = "cv")
```

The `label_style` argument of `plot_triangle()` switches the cell
labels between value (`"value"`), coefficient of variation (`"cv"`),
standard error (`"se"`), and confidence interval (`"ci"`). For the
detailed arguments of `plot()` -- `type`, `region`, and so on -- see
each class's help page (`?plot.CLFit` and friends).

The CL / ED / SA / BF / CC / premium fit results all support `plot()`
and `plot_triangle()` the same way, so the visualisation interface is
identical regardless of method.

## See also

This tutorial follows the core workflow in one pass. For the
methodological background and detailed options of each topic, see:

- `?as_triangle`, `?fit_cl`, `?fit_sa`, `?fit_ratio` and other
  per-function help pages.
- `vignette("projection")` -- projection methods in depth.
- `vignette("diagnostics")` -- maturity, convergence, and regime
  detection.
- `vignette("backtest")` -- the validation procedure.
- `vignette("bootstrap-se-decomposition")` -- bootstrap standard
  errors and variance decomposition.
</content>
</invoke>
