---
title: "Detecting LR stability with find_lr_stability()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting LR stability with find_lr_stability()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

## Motivation

`find_ata_maturity()` answers the question *"from which development
period are link factors $f_k$ reproducible across cohorts?"*. That is
necessary for chain-ladder projection but not sufficient for declaring
a portfolio's loss ratio stable: in long-duration health insurance both
$f_k \to 1$ and $g_k \to 0$ arise mechanically from cumulative
denominators growing, regardless of whether the underlying experience
has actually settled. A criterion built on those quantities passes
automatically with $k$, not because of stability — what we have called
the *inertia* failure mode.

`find_lr_stability()` implements the dual criterion from the paper's
Section 11 ($k_0$). It detects the first valuation $v$ at which **two**
conditions hold for $M$ consecutive periods:

1. **Predictive revision** is small relative to its parameter SE:
   $R_v < c \cdot \widehat{SE}^{\mathrm{param}}_v$, where
   $R_v = |\widehat{LR}^{\mathrm{proj}}_v(D_v) -
   \widehat{LR}^{\mathrm{proj}}_v(D_{v-1})|$ is the change in the
   projected portfolio LR caused by adding one new calendar diagonal.
2. **Cross-cohort dispersion** of incremental LR is small:
   $\widehat{D}_v < \tau$, where
   $\widehat{D}_v = 1.4826 \cdot \mathrm{MAD}_i(\hat{lr}_{i,v}) /
   |\mathrm{median}_i(\hat{lr}_{i,v})|$.

Operating $\widehat{D}_v$ on incremental rather than cumulative loss
ratio keeps it inertia-free — per-period quantities have no
cumulative denominator to dampen them. The pair guards against
complementary failure modes: $R_v$ checks that the model has stopped
learning, and $\widehat{D}_v$ checks that the underlying experience is
genuinely consistent across cohorts at that dev.

## Basic usage

```{r, eval = FALSE}
library(lossratio)
data(experience)
exp <- as_experience(experience)
tri <- build_triangle(exp[cv_nm == "SUR"], cv_nm)

res <- find_lr_stability(tri)
print(res)
```

Mock output:

```
#> <LRStability>
#> k_stable     : NA
#> k_star       : 9
#> V (max dev)  : 30
#> criterion    : R_v < 0.5 * SE_param_v  AND  D_v < 0.15  (run M = 3)
#> fit_fn       : fit_lr
#> v candidates : 19 ( 0  pass both clauses)
```

The returned `LRStability` object reports:

- `k_stable` — the detected $k_0$, or `NA` if no run of $M$
  consecutive passing valuations is found.
- `k_star` — the maturity point used as the lower bound (computed
  internally via `find_ata_maturity()` on a clr-based ATA, or supplied
  by the caller).
- `V` — the maximum observable dev in the triangle.
- `v`, `R_v`, `SE_param_v`, `D_v`, `pass_v` — per-valuation diagnostic
  sequences indexed by $v$.
- `c`, `tau`, `M`, `holdout_max`, `min_n_cohorts` — settings used.
- attributes `group_var`, `value_var`, `fit_fn_name`, `dev_var`.

`summary(res)` returns a `data.table` with one row per candidate
valuation and an extra `R_over_SE = R_v / SE_param_v` column for
inspection:

```{r, eval = FALSE}
head(summary(res), 6)
```

```
#>        v    R_v   SE_param_v  R_over_SE   D_v     pass
#> 1:     9     NA           NA         NA  0.90   FALSE
#> 2:    10     NA           NA         NA  0.76   FALSE
#> 3:    11     NA           NA         NA  0.56   FALSE
#> 4:    12     NA           NA         NA  0.58   FALSE
#> 5:    13     NA           NA         NA  0.81   FALSE
#> 6:    14     NA           NA         NA  0.43   FALSE
```

`R_v` and `SE_param_v` are `NA` for early valuations whose holdout
depth exceeds `holdout_max`. The default `holdout_max = floor((V -
k_star) / 2)` keeps every backtest with a reasonable refit window;
relax it if you need diagnostics deeper into the past.

## Plot

```{r, eval = FALSE}
plot(res)
```

The diagnostic is two stacked panels: the upper panel shows
$R_v / \widehat{SE}^{\mathrm{param}}_v$ against $v$ with a horizontal
guide at $c$; the lower panel shows $\widehat{D}_v$ against $v$ with a
horizontal guide at $\tau$. A vertical dotted line marks $k^*$, and a
vertical solid line marks $k_{\mathrm{stable}}$ when one is detected.
A point falling **below both threshold lines** passes the joint
criterion.

This view is also a quick way to see *which clause is binding*. If the
top panel hugs the threshold but the bottom is far above, the issue is
cross-cohort heterogeneity; if the bottom is fine but the top is high,
the model is still revising.

## Threshold tuning

The defaults are deliberately conservative:

| Argument | Default | Meaning |
|----------|---------|---------|
| `c`      | `0.5`   | Revision must be smaller than half the parameter SE. |
| `tau`    | `0.15`  | Cross-cohort dispersion must be below 15% of the median lr. |
| `M`      | `3L`    | Both clauses must hold for at least 3 consecutive valuations. |
| `min_n_cohorts` | `5L` | Below this cohort count, $\widehat{D}_v$ is `NA` (insufficient sample). |

Tighter thresholds yield later (or no) $k_{\mathrm{stable}}$; sweep a
range to inspect sensitivity:

```{r, eval = FALSE}
sapply(
  c(0.25, 0.5, 0.75, 1.0),
  function(cc) find_lr_stability(tri, c = cc)$k_stable
)
```

Values of $\widehat{D}_v$ below $\tau \approx 0.05$ are difficult to
attain in real portfolios because of single-period claim noise; values
above $0.20$ usually indicate genuine cohort heterogeneity that
warrants `detect_cohort_regime()` before further modelling.

## Relation to `k^*` and `detect_cohort_regime()`

The three diagnostics answer different questions and operate on
different axes:

| Tool | Question | Result | Axis |
|------|----------|--------|------|
| `detect_cohort_regime()` | Are cohorts homogeneous? | cohort groups | underwriting period |
| `find_ata_maturity()` ($k^*$) | When are link factors reproducible? | a dev value | development period |
| `find_lr_stability()` ($k_0$) | When does the LR estimate stop revising? | a dev value | development period |

A defensible workflow is:

1. Run `detect_cohort_regime()`. If multiple regimes exist, fit each
   group separately.
2. For each homogeneous group, compute $k^*$ via
   `find_ata_maturity()`.
3. Run `find_lr_stability()` to obtain $k_0 \ge k^*$. The reported
   $\widehat{clr}^{\mathrm{stable}}$ is the LR averaged over
   $k \ge k_0$ (or projected via `fit_lr()`).

This sequence reflects the paper's structural separation of *cohort
homogeneity*, *link reproducibility*, and *level convergence* — three
properties that coincide in P&C run-off but must be verified
independently in long-duration health insurance.

## Limitations

`find_lr_stability()` is a thin layer over repeated `backtest()` calls
and inherits their constraints:

- **Identifiability**: $k_0$ can be declared only when $V \ge k^* + M$;
  short observation windows return `NA`.
- **Model conditioning**: $\widehat{LR}^{\mathrm{proj}}_v$ is computed
  by `fit_fn` (default `fit_lr`). Different fitters yield different
  $k_0$. Reporting under multiple `fit_fn` is recommended for
  robustness.
- **Portfolio aggregation**: $R_v$ and $\widehat{SE}^{\mathrm{param}}_v$
  are exposure-weighted across cohorts assuming inter-cohort
  independence. Calendar-year shocks (regulatory, healthcare cost
  trend) violate this assumption; both clauses can move together for
  non-cohort reasons.
- **Multi-group triangles**: the helper currently collapses
  $\widehat{D}_v$ across groups by median; running each group
  separately is recommended when groups behave differently.

## See also

- `?find_lr_stability`, `?find_ata_maturity`, `?backtest`,
  `?detect_cohort_regime`.
- Companion paper: *From Exposure to Experience: A Stage-Adaptive
  Chain Ladder for Long-Duration Health Insurance*, Section 11
  ("Determination of $k_0$").
