---
title: "Maturity: detecting when ATA factors stabilise across cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maturity: detecting when ATA factors stabilise across cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

The maturity point is the development link beyond which age-to-age
factors are stable enough to trust for chain-ladder projection. Used
internally by `fit_lr(method = "sa")` to switch from ED to CL. The
factor diagnostics that drive detection are covered in
`vignette("triangle-and-link")`; this article focuses on
`detect_maturity()` itself.

## Setup

For brevity this vignette uses the `SUR` group only — every step
generalises to multi-group input.

```{r}
library(lossratio)
data(experience)
exp <- as_experience(experience)[cv_nm == "SUR"]
tri <- build_triangle(exp, group_var = cv_nm)
```

## Detecting maturity

`detect_maturity()` takes a `Triangle` directly — the underlying
single-variable `Link` and its WLS summary are built internally:

```{r}
mat <- detect_maturity(
  tri,
  value_var       = "closs",
  cv_threshold    = 0.10,    # CV must be below this
  rse_threshold   = 0.05,    # RSE must be below this
  min_valid_ratio = 0.5,     # at least 50% finite cohorts at the link
  min_n_valid     = 3L,      # at least 3 finite cohorts
  min_run         = 1L       # at least 1 consecutive mature link
)

print(mat)
```

A row per group with the first development link satisfying all
thresholds, carrying the link's full statistics. The threshold
arguments are also stored as attributes on the returned `Maturity`
object.

## Threshold semantics

- `cv_threshold` — coefficient of variation of the observed ATA
  factors at the link. Caps relative spread regardless of `alpha`.
- `rse_threshold` — relative standard error of the WLS-estimated
  factor `f`. Captures parameter uncertainty rather than residual
  spread.
- `min_valid_ratio` — minimum share of cohorts with a finite ATA at
  the link. Guards against links where most observations are zero
  / NA / Inf.
- `min_n_valid` — minimum count of finite cohorts at the link.
  Absolute floor for thin-data tails.
- `min_run` — minimum number of *consecutive* mature links. With
  `min_run = 1L` (default) the first qualifying link wins; setting
  it to `2L` or higher requires sustained stability.

Tune these to your portfolio's volatility profile. Tight thresholds
(e.g. `cv_threshold = 0.05`) push maturity later; loose thresholds
push it earlier.

## Use in fitting

`detect_maturity()` is also called internally by `fit_ata()` and
`fit_cl()` when `maturity_args` is supplied (the `alpha` of the
internal `summary()` step is taken from those callers):

```{r, eval = FALSE}
fit_ata(tri, value_var = "closs",
        maturity_args = list(cv_threshold = 0.08, min_run = 2L))

fit_cl(tri, value_var = "closs",
       maturity_args = list(cv_threshold = 0.08))

fit_lr(tri, method = "sa",
       maturity_args = list(cv_threshold = 0.08))
```

For `fit_lr(method = "sa")` the detected maturity point determines
the dev at which the projection switches from ED (early dev) to CL
(later dev).

## Group-wise output

For multi-group triangles `detect_maturity()` returns one row per
group:

```{r, eval = FALSE}
tri_all <- build_triangle(as_experience(experience), group_var = cv_nm)
detect_maturity(tri_all, value_var = "closs")
```

Each group is detected independently with the same thresholds.

## See also

- `vignette("triangle-and-link")` — `Triangle` / `Link` data
  structures and the per-link statistics used by `detect_maturity()`.
- `vignette("loss-ratio-methods")` — how the maturity point feeds
  into `fit_lr(method = "sa")`.
- `?detect_maturity` — full argument reference.
