---
title: "Projection: SA, ED, CL methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projection: SA, ED, CL methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

`fit_ratio()` projects cumulative loss ratio per cohort from a
`Triangle` object. Three methods are available; this vignette
explains the trade-offs.

## Notation

For cohort $i$ at dev $k$:

- $C^L_{i,k}$ — cumulative loss
- $C^P_{i,k}$ — cumulative risk premium (premium)
- $f_k = C^L_{k+1} / C^L_k$ — age-to-age (chain ladder) factor
- $g_k = \Delta C^L_k / C^P_k$ — exposure-driven intensity
- maturity point $k^*$ — dev at which $f_k$ stabilises
  for group $g$ (detected from CV / RSE thresholds)

## Method 1: Exposure-Driven (`"ed"`, default)

All future increments use premium (risk premium) as the
denominator:

$$
\hat{C}^L_{i,k+1} = \hat{C}^L_{i,k} + g_k \cdot C^P_{i,k}
$$

ED is the default because it is an *unconditional safe baseline*
-- no maturity or regime detection is required, and the projection
is robust under early-dev age-to-age volatility.

Behaviour:

- Stable when premium volume is informative across full development.
- The pooled intensity $g_k$ assumes cohorts are reasonably
  homogeneous in loss-per-premium level. Under cohort-level drift
  the projection biases toward the pooled mean and may over-project
  post-change cohorts -- see the `regime` argument for explicit
  filtering.

When to use:

- As a baseline projection where cohort homogeneity is plausible.
- Short-tail products where chain ladder offers no advantage.
- Sparse data where age-to-age factors are unreliable across all
  links.

```{r}
library(lossratio)
data(experience)
tri <- as_triangle(
  experience[coverage == "surgery"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_premium"
)

ratio_ed <- fit_ratio(tri, method = "ed")        # default
plot(ratio_ed, metric = "ratio")
summary(ratio_ed)
```

## Method 2: Classical Chain Ladder (`"cl"`)

Classical Mack (1993) model:

$$
\hat{C}^L_{i,k+1} = f_k \cdot \hat{C}^L_{i,k}
$$

The cohort's own cum_loss acts as the anchor, so cohort-level drift
propagates naturally without explicit regime detection.

Behaviour:

- Standard reserving practice. Equivalent to `fit_cl(tri, loss = "loss")`
  for the loss projection, but `fit_ratio()` additionally projects
  premium forward via CL on `premium` and computes the loss-ratio
  uncertainty via the delta method.
- Volatile when early $f_k$ are noisy -- small denominators amplify
  link errors.

When to use:

- Once age-to-age factors stabilise.
- Cohort-level drift scenarios where the cohort's observed
  trajectory should anchor the projection.
- Reserving exercises where regulators expect the classical Mack
  form for documentation.

```{r}
ratio_cl <- fit_ratio(tri, method = "cl")
plot(ratio_cl, metric = "ratio")
```

## Method 3: Stage-Adaptive (`"sa"`)

SA composes ED before the maturity point with CL after, exploiting
the fact that $f_k$ is volatile early and stable late, while $g_k$
behaves the opposite way:

$$
\hat{C}^L_{i,k+1} \;=\;
\begin{cases}
\hat{C}^L_{i,k} + g_k \cdot C^P_{i,k} & k < k^* \quad \text{(ED before maturity)} \\
f_k \cdot \hat{C}^L_{i,k}              & k \ge k^* \quad \text{(CL after maturity)}
\end{cases}
$$

Behaviour:

- **Before maturity**: anchors the loss estimate to premium volume.
  Avoids the volatile-link explosion that classical CL suffers when
  early $f_k$ are noisy.
- **After maturity**: preserves the cohort's own observed level.
  Avoids the "all cohorts converge to the average" behaviour that
  pure ED suffers in the tail.

When to use:

- Long-tail portfolios where early dev is volatile (ED phase) and
  cohort-level drift needs cohort-anchored projection in later dev
  (CL phase).
- Recent cohorts (immature data) mixed with older cohorts (matured).
- Health insurance cohorts with structural pre-/post-maturity
  difference (e.g. waiting period transitions).

```{r}
ratio_sa <- fit_ratio(tri, method = "sa")
plot(ratio_sa, metric = "ratio")
summary(ratio_sa)
```

## Method summary

| method | mechanism                       | when to use                                                         |
|--------|---------------------------------|---------------------------------------------------------------------|
| ED     | pooled $g_k$ x cohort premium  | default baseline; assumes cohort homogeneity                        |
| CL     | pooled $f_k$ x cohort cum_loss  | cohort-level drift; cohort-anchored                                 |
| SA     | ED early + CL late              | long-tail with both volatile early dev and cohort-level late drift  |

## Comparison

```{r}
ratios <- list(
  ed = fit_ratio(tri, method = "ed"),
  cl = fit_ratio(tri, method = "cl"),
  sa = fit_ratio(tri, method = "sa")
)

# Cohort-level summary
summary(ratios$ed)$loss_ult
summary(ratios$cl)$loss_ult
summary(ratios$sa)$loss_ult
```

## Variance and confidence intervals

`fit_ratio()` reports analytical standard errors for `L/P`. Two variants:

- `se_method = "fixed"` (default) — premium treated as fixed,
  $\mathrm{SE}(L/P) \approx \mathrm{SE}(L)/P$. Strictly the delta method
  with `Var(P) = 0` and `Cov(L,P) = 0`.
- `se_method = "delta"` — full delta method including premium uncertainty
  and loss-premium correlation `rho`:

$$
\mathrm{Var}(L/P) \approx \frac{\mathrm{Var}(L)}{P^2}
  + \frac{L^2 \mathrm{Var}(P)}{P^4}
  - \frac{2 \rho L \mathrm{SE}(L) \mathrm{SE}(P)}{P^3}
$$

Bootstrap intervals are also available:

```{r}
ratio_boot <- fit_ratio(tri, method = "ed", bootstrap = TRUE, B = 1000, seed = 1)
summary(ratio_boot)
```

## Choosing a method

ED is the default because it is the *unconditional safe baseline*:
no maturity or regime detection is required, and it is conservative
under early-dev volatility. CL and SA are warranted when the
default's homogeneity assumption is unrealistic.

```
Default is "ed" -- pooled g_k x cohort premium, no maturity dependency.

Move to "cl" or "sa" when ED's assumptions become limiting:
  |-- Cohort-level drift (regime change, underwriting / coverage shift)
  |     -> "cl"  (cohort's own cum_loss anchors the drift)
  |-- Long-tail portfolio with BOTH volatile early dev AND cohort drift late
  |     -> "sa"  (ED before maturity, CL after)
  |-- All cohorts already past maturity, no early-dev volatility
        -> "cl"  (no ED region needed)
```

In practice: **start with `"ed"`** (the default), then run `"cl"`
and `"sa"` for sensitivity. If all three agree, the projection is
robust. If they diverge, inspect maturity detection, regime filters,
and the underlying ATA factors.

## Maturity input

For `method = "sa"` the maturity point $k^*$ determines where the
projection switches from ED to CL. By default `fit_ratio()` /
`fit_loss()` use `maturity = "auto"`, which calls
`detect_maturity()` internally with default thresholds. Other forms:

- `maturity = NULL` — disable detection (only meaningful for
  non-SA methods or for `fit_ata()` standalone).
- `maturity = maturity_spec(max_cv = 0.05, min_run = 2L)` — lazy
  spec forwarding custom thresholds to `detect_maturity()` at fit
  time (leakage-safe for `backtest()`).
- `maturity = detect_maturity(tri, ...)` — pre-built `Maturity`
  object, fixed across refits.
- `maturity = maturity_at(coverage = "surgery", change = 4)` — manual
  per-group override (e.g. company-standard $k^*$).
