---
title: "Projection methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projection methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

> Korean version: [예측 방법론](articles/projection-ko.html)

This is a deep-dive into the five projection methods in `lossratio` —
exposure-driven (ED), chain ladder (CL), stage-adaptive (SA),
Bornhuetter-Ferguson (BF), and Cape Cod (CC). The getting-started
tutorial shows *how* to call each fit; this document explains *why*
each method exists, what assumption it rests on, and when one is
preferable to another. The intended reader is a practitioner working
on long-term health insurance loss ratios.

## Why a dedicated projection toolkit

Loss ratio is a daily task in long-term health insurance: analysing
cohort-level patterns, projecting ultimate outcomes, and monitoring
realised experience against expectations. Existing reserving toolkits
mostly trace back to property and casualty (P&C) origins, where the
following long-term health characteristics are not directly captured:

- **Denominator effect and inertia** — Cumulative loss ratio's
  denominator grows mechanically with development, dampening the
  signal of recent experience.
- **Recurrent claims** — Hospitalisation, surgery (grade 1-5), and
  outpatient coverage allow a single insured to file multiple claims;
  cumulative claim count can exceed insured count.
- **Risk premium decomposition** — In Korea, long-term insurance
  premium splits into risk premium + savings premium + loading
  premium. The actuarially meaningful denominator for loss ratio is
  the *risk premium* portion alone.
- **Levelled premium** — Non-renewable contracts charge a level
  premium computed at issue as the lifetime average. The *charged
  premium* is not premium; the *period risk premium*
  (morbidity rate x sum insured x persistency) must be constructed
  externally before being passed in.
- **Regime change** — Product redesigns, rate revisions, channel-mix
  shifts, and underwriting updates introduce *cohort-level*
  structural breaks in loss-ratio dynamics.

`lossratio` transplants core P&C reserving methodology — Mack (1993)
chain ladder, Bornhuetter-Ferguson (1972), Cape Cod (Stanard 1985),
Bühlmann-Straub (1970) credibility, and Sherman (1984) tail
extrapolation — and adapts each piece to the long-term health issues
above. The academic lineage is preserved; the goal is a tool that
domain practitioners can use immediately.

## Methodological lineage

The methodological roots of long-term loss-ratio estimation lie in the
following P&C reserving thread:

```
1967  Bühlmann credibility           experience rating formalised
1970  Bühlmann-Straub                premium-varying credibility
                                     (volume-weighted estimator)
1972  Bornhuetter-Ferguson           prior + observed loss blending
1984  Sherman                        chain ladder tail extrapolation
1985  Stanard (Cape Cod)             reserving application of B-S
1993  Mack                           distribution-free MSE for chain
                                     ladder
```

Two core ideas drive everything that follows:

- **Chain ladder (Mack 1993)** — Markov multiplicative recursion on
  cumulative loss \(C_k\): \(C_{k+1} = f_k \cdot C_k\) where
  \(f_k = \sum_i C_{i,k+1} / \sum_i C_{i,k}\).
- **Cape Cod / Bühlmann-Straub** — Loss anchored to premium (volume).
  \(\widehat{ELR} = \sum_i L_i / \sum_i \pi_i\) yields a single ratio
  that drives ultimate estimation.

Both apply *partially* to long-term health, but both have cracks:

| Domain issue | Chain ladder | Cape Cod |
|---|---|---|
| Denominator effect / inertia | Early-dev \(f_k\) over-volatile (small \(C_k\)) | Cohort-level ELR variation ignored |
| Levelled premium | Loss-only avoids it (weak against incidence shifts) | Flat \(\pi\) absorbs developing premium |
| Recurrent claims | Works (freq x sev decoupled) | Works (volume measure only) |
| Developing risk premium | Not relevant | *Single \(\pi\)* assumption breaks |
| Cohort-level regime change | Violates Mack's no-calendar-year-effect | Cohort heterogeneity averaged out |

Chain ladder is weak early; Cape Cod cannot track developing premium.
Each paradigm is incomplete on its own — which is exactly why the
package offers five methods rather than one.

## The core framework: loss / premium / ratio

All estimation rests on three observable quantities:

| Quantity | Meaning | Triangle column |
|---|---|---|
| **loss** | Cumulative loss | `loss`, `incr_loss` |
| **premium** | Risk-bearing volume (risk premium for long-term health) | `premium`, `incr_premium` |
| **ratio** | Loss ratio (cumulative loss / cumulative premium) | `ratio`, `incr_ratio` |

All three are *stochastic observables developing over the cohort x dev
grid*. Premium is not a fixed underwriting volume; it is a developing
quantity driven by morbidity x sum insured x persistency — the *volume
measure* of Mack (1993), the *natural weight* of Bühlmann-Straub
(1970).

For cohort \(i\) at development period \(k\) the projection methods use
this notation:

- \(C^L_{i,k}\) — cumulative loss
- \(C^P_{i,k}\) — cumulative risk premium (premium)
- \(f_k = C^L_{k+1} / C^L_k\) — age-to-age (chain ladder) factor
- \(g_k = \Delta C^L_k / C^P_k\) — exposure-driven intensity
- maturity point \(k^*\) — the development period at which \(f_k\)
  stabilises for a group (detected from CV / RSE thresholds)

We use the `surgery` group throughout for brevity — every step
generalises to multi-group input.

```{r}
library(lossratio)
data(experience)
tri <- as_triangle(
  experience[coverage == "surgery"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_premium"
)
```

## Direct estimation: ED, CL, SA

The three direct-estimation methods estimate factors from the data and
chain them forward. Their order — `ed` -> `cl` -> `sa` — mirrors the
methodological progression *primitive (ED) -> classical (CL) ->
composition (SA)*.

| Method | Point estimate | Variance helper | Domain character |
|---|---|---|---|
| `"ed"` (default) | \(\Delta L_{k+1} = g_k \cdot P_k\) (additive) | `.ed_g_var` (B-S 1970) | Robust to early-dev ATA volatility |
| `"cl"` | \(L_{k+1} = f_k \cdot L_k\) (multiplicative) | `.mack_f_var` (Mack 1993) | Natural after late-dev factor stabilisation |
| `"sa"` | ED before maturity \(k^*\), CL after | Composition | Stage-adaptive composition of ED and CL |

### Exposure-driven (`"ed"`, default)

ED projects every future increment using premium (risk premium) as the
denominator:

$$
\hat{C}^L_{i,k+1} = \hat{C}^L_{i,k} + g_k \cdot C^P_{i,k}
$$

ED is the default because it is an *unconditional safe baseline*: no
maturity or regime detection is required, and the projection is robust
under early-dev age-to-age volatility. Because the small \(C_k\) of
early dev never enters the denominator, ED does not inflate the
estimate the way an early \(f_k\) can.

The trade-off is cohort homogeneity. The pooled intensity \(g_k\)
assumes cohorts are reasonably homogeneous in loss-per-premium level.
Under cohort-level drift the projection biases toward the pooled mean
and may over-project post-change cohorts — see the `regime` argument
for explicit filtering.

**When to use ED:** as a baseline where cohort homogeneity is
plausible; short-tail products where chain ladder offers no advantage;
sparse data where age-to-age factors are unreliable across all links.

```{r}
ratio_ed <- fit_ratio(tri, method = "ed")        # default
plot(ratio_ed, metric = "ratio")
summary(ratio_ed)
```

### Classical chain ladder (`"cl"`)

CL is the classical Mack (1993) model:

$$
\hat{C}^L_{i,k+1} = f_k \cdot \hat{C}^L_{i,k}
$$

The cohort's own cumulative loss acts as the anchor, so cohort-level
drift propagates naturally without explicit regime detection. The
trade-off is the mirror image of ED's: CL is volatile when early
\(f_k\) are noisy, because small denominators amplify link errors.

Within `fit_ratio()`, the CL method projects loss *and* premium
forward — each via chain ladder on its own column — and computes the
loss-ratio uncertainty via the delta method. The loss lane alone is
equivalent to `fit_cl(tri)`.

**When to use CL:** once age-to-age factors stabilise; cohort-level
drift scenarios where the cohort's observed trajectory should anchor
the projection; reserving exercises where regulators expect the
classical Mack form for documentation.

```{r}
ratio_cl <- fit_ratio(tri, method = "cl")
plot(ratio_cl, metric = "ratio")
```

### Stage-adaptive (`"sa"`)

SA composes ED before the maturity point with CL after, exploiting the
fact that \(f_k\) is volatile early and stable late, while \(g_k\)
behaves the opposite way:

$$
\hat{C}^L_{i,k+1} \;=\;
\begin{cases}
\hat{C}^L_{i,k} + g_k \cdot C^P_{i,k} & k < k^* \quad \text{(ED before maturity)} \\
f_k \cdot \hat{C}^L_{i,k}              & k \ge k^* \quad \text{(CL after maturity)}
\end{cases}
$$

- **Before maturity** SA anchors the loss estimate to premium volume,
  avoiding the volatile-link explosion that classical CL suffers when
  early \(f_k\) are noisy.
- **After maturity** SA preserves the cohort's own observed level,
  avoiding the "all cohorts converge to the average" behaviour that
  pure ED suffers in the tail.

**When to use SA:** long-tail portfolios where early dev is volatile
(ED phase) and cohort-level drift needs cohort-anchored projection in
later dev (CL phase); recent cohorts (immature data) mixed with older
cohorts (matured); health insurance cohorts with a structural
pre-/post-maturity difference (e.g. waiting-period transitions).

```{r}
ratio_sa <- fit_ratio(tri, method = "sa")
plot(ratio_sa, metric = "ratio")
summary(ratio_sa)
```

### Comparing ED, CL, SA

```{r}
ratios <- list(
  ed = fit_ratio(tri, method = "ed"),
  cl = fit_ratio(tri, method = "cl"),
  sa = fit_ratio(tri, method = "sa")
)

# Cohort-level ultimate-loss summary
summary(ratios$ed)$loss_ult
summary(ratios$cl)$loss_ult
summary(ratios$sa)$loss_ult
```

| Method | Mechanism | When to use |
|---|---|---|
| ED | pooled \(g_k\) x cohort premium | default baseline; assumes cohort homogeneity |
| CL | pooled \(f_k\) x cohort cum_loss | cohort-level drift; cohort-anchored |
| SA | ED early + CL late | long-tail with both volatile early dev and cohort-level late drift |

## Prior-anchored estimation: BF and CC

ED, CL, and SA all estimate factors *from the data*. When the observed
triangle is thin — immature cohorts, or cohorts right after a rate
change — that estimate can be unstable. The prior-anchored family
blends an expected loss ratio (ELR) with whatever loss has already
emerged:

$$
\text{Ult} = L_{\text{latest}}
  + \left(1 - \tfrac{1}{\text{LDF}}\right) \cdot \pi \cdot \text{ELR}
$$

The two methods differ only in *where the ELR comes from*.

| Method | ELR source | Domain use case |
|---|---|---|
| `"bf"` | External (user supplied via `prior`) | Immature and post-rate-change cohorts — anchor on an external prior when observed data is thin (Bornhuetter-Ferguson 1972) |
| `"cc"` | Derived from data (payout-weighted \(\sum L / \sum \pi \cdot \text{payout}\)) | Cohort-cohesive estimation — when pricing/industry suggests a natural single ELR target (Stanard 1985, Cape Cod) |

### Bornhuetter-Ferguson (`"bf"`)

BF takes the ELR as an external input. The emerged loss
\(L_{\text{latest}}\) is kept as-is, and only the *unemerged* portion
\((1 - 1/\text{LDF})\) is filled in from the prior. As a cohort
matures the data term dominates and the prior fades — BF degrades
gracefully toward a chain-ladder answer.

```{r}
fit_bf(tri, prior = 0.7)
```

The `prior` argument also accepts a per-group `data.frame`, optionally
carrying an `elr_se` column to treat the prior as a *distribution*
rather than a fixed point.

### Cape Cod (`"cc"`)

CC has the same blending form as BF but estimates the ELR *from the
data itself* — a payout-weighted pooled ratio across cohorts. It is
the natural choice when pricing or industry context suggests a single
coherent ELR target and no external prior is on hand.

```{r}
fit_cc(tri)
```

### Three aggregation axes

ED, CC, and BF differ in how they aggregate the loss-per-premium
signal:

- **ED** — per-link \(g_k\): dev-granular, cohort-uniform per link.
- **CC** — cohort-level single ELR: cohort-uniform, dev-aggregated.
- **BF** — external prior: bypasses data estimation altogether.

Together the five methods reconstruct the P&C reserving trinity for
long-term health, and the three aggregation axes each serve a distinct
use case.

## Chain ladder as a reserving worker

`fit_cl()` is the dedicated chain ladder fit for a single value
column. Unlike `fit_ratio()` — which projects loss and premium jointly
to get loss ratio — `fit_cl()` projects one cumulative metric forward
and computes Mack-style standard errors per cohort. This is the
classical P&C *reserving* use case: projecting ultimate paid /
incurred loss for an open accident year. Practitioners with a P&C
background will recognise the Mack workflow directly.

```{r}
cl <- fit_cl(tri)
print(cl)
```

`fit_cl()` summarises adjacent development links by age-to-age factors
\(f_k = C^L_{k+1} / C^L_k\), selected per link and then chained to
project each cohort forward to ultimate. On top of the point
projection, Mack's formulae decompose the prediction variance into
process and parameter components:

- `loss_proc_se` — process variance, from \(\sigma^2_k\) (residual
  link variance per development period).
- `loss_param_se` — parameter variance, from the uncertainty of the
  selected age-to-age factors \(\hat{f}_k\).
- `loss_total_se` — total standard error,
  \(\sqrt{\text{loss\_proc\_se}^2 + \text{loss\_param\_se}^2}\).
- `loss_total_cv` — coefficient of variation,
  `loss_total_se / loss_proj`.

```{r}
summary(cl)
plot(cl, type = "projection", show_interval = TRUE)
plot(cl, type = "reserve", conf_level = 0.95)
```

`plot_triangle()` displays the cohort x dev cells as a heatmap,
distinguishing observed cells from projected, with `label_style`
showing per-cell CV / SE / CI:

```{r}
plot_triangle(cl, region = "full")        # observed + projected
plot_triangle(cl, label_style = "cv")     # per-cell coefficient of variation
```

### Tail factor

For triangles where the latest observed development period is still
developing, an extrapolated tail factor estimates ultimate. The
extrapolation fits \(\log(f_k - 1) \sim k\) — a Sherman (1984)
log-linear tail — to the selected ATA factors and extends the
projection by the cumulative product of extrapolated \(f_k\) values.
Disabled by default (`tail = FALSE`).

```{r}
fit_cl(tri, tail = TRUE)         # log-linear extrapolation
fit_cl(tri, tail = 1.025)        # or a literal tail factor
```

### Sigma extrapolation methods

Mack variance requires \(\sigma_k\) at all development links,
including the last where it cannot be estimated directly.
`sigma_method` controls the extrapolation:

| `sigma_method` | Behaviour |
|---|---|
| `"locf"` | (default) last observation carried forward |
| `"min_last2"` | min of the last two estimable \(\sigma\) values — conservative |
| `"loglinear"` | log-linear extrapolation from the observed \(\sigma_k\) sequence |
| `"mack"` | Mack (1993) Appendix B tail estimator — closed-form for the last link only; LOCF with warning beyond that |
| `"none"` | no extrapolation; \(\sigma\) stays `NA` (variance treated as zero downstream) |

```{r}
fit_cl(tri, sigma_method = "loglinear")
```

## Variance and confidence intervals

`fit_ratio()` reports analytical standard errors for `L/P`. Two
variants control how premium uncertainty enters:

- `se_method = "fixed"` (default) — premium treated as fixed,
  \(\text{SE}(L/P) \approx \text{SE}(L)/P\). Strictly the delta method
  with `Var(P) = 0` and `Cov(L,P) = 0`.
- `se_method = "delta"` — full delta method including premium
  uncertainty and loss-premium correlation `rho`:

$$
\mathrm{Var}(L/P) \approx \frac{\mathrm{Var}(L)}{P^2}
  + \frac{L^2 \mathrm{Var}(P)}{P^4}
  - \frac{2 \rho L \mathrm{SE}(L) \mathrm{SE}(P)}{P^3}
$$

There are two complementary paths to SE estimation:

- **Analytical** — `.mack_f_var()` (Mack 1993) and `.ed_g_var()`
  (B-S 1970) provide closed-form per-link variance, distribution-free.
- **Bootstrap** — a residual paradigm (`cell` / `link` / `parametric`)
  drives forward simulation, capturing distributional shape. Residual
  choice determines both the process variance scale and the
  forward-sim model (paradigm matching).

A large divergence between the two paths is a model-misspecification
signal; routinely computing both serves as a sanity check.

```{r}
ratio_boot <- fit_ratio(tri, method = "ed", bootstrap = TRUE,
                        B = 1000, seed = 1)
summary(ratio_boot)
```

## Maturity input for SA

For `method = "sa"` the maturity point \(k^*\) determines where the
projection switches from ED to CL. By default `fit_ratio()` /
`fit_loss()` use `maturity = "auto"`, which calls `detect_maturity()`
internally with default thresholds. Other forms:

- `maturity = NULL` — disable detection (only meaningful for non-SA
  methods or for `fit_ata()` standalone).
- `maturity = maturity_spec(max_cv = 0.05, min_run = 2L)` — a lazy
  spec forwarding custom thresholds to `detect_maturity()` at fit time
  (leakage-safe for `backtest()`).
- `maturity = detect_maturity(tri, ...)` — a pre-built `Maturity`
  object, fixed across refits.
- `maturity = maturity_at(coverage = "surgery", change = 4)` — a
  manual per-group override (e.g. a company-standard \(k^*\)).

The same lazy-spec / pre-built / manual pattern applies to `fit_cl()`,
where maturity filtering restricts the projection to the mature
region when the selected ATA factors are volatile.

## Choosing a method

ED is the default because it is the *unconditional safe baseline*: no
maturity or regime detection is required, and it is conservative under
early-dev volatility. The other methods are warranted when ED's
homogeneity assumption is unrealistic, or when an external anchor is
preferable to factors estimated from a thin triangle.

```
Default is "ed" -- pooled g_k x cohort premium, no maturity dependency.

Move to another method when ED's assumptions become limiting:
  |-- Cohort-level drift (regime change, underwriting / coverage shift)
  |     -> "cl"  (cohort's own cum_loss anchors the drift)
  |-- Long-tail portfolio with BOTH volatile early dev AND cohort drift
  |     -> "sa"  (ED before maturity, CL after)
  |-- All cohorts already past maturity, no early-dev volatility
  |     -> "cl"  (no ED region needed)
  |-- Immature / post-rate-change cohorts, an external ELR is on hand
  |     -> "bf"  (blend the external prior with emerged loss)
  |-- A single coherent ELR target, no external prior available
        -> "cc"  (data-pooled ELR, same blending form as BF)
```

In practice: **start with `"ed"`** (the default), then run `"cl"` and
`"sa"` for sensitivity. If all three agree, the projection is robust.
If they diverge, inspect maturity detection, regime filters, and the
underlying ATA factors. Reach for `"bf"` / `"cc"` when the cohort is
too immature for any data-only method to be trustworthy.

## See also

- `vignette("getting-started")` — package quick-start guide.
- `vignette("diagnostics")` — regime detection and filtering.
- `vignette("backtest")` — calendar-diagonal hold-out validation.
- `vignette("diagnostics")` — projected loss-ratio convergence
  diagnostic.
- `vignette("getting-started")` — Triangle / Link /
  Maturity data flow.

## References

- Bornhuetter, R. L. and Ferguson, R. E. (1972). The actuary and IBNR.
  *Proceedings of the Casualty Actuarial Society*, 59, 181-195.
- Bühlmann, H. (1967). Experience rating and credibility. *ASTIN
  Bulletin*, 4(3), 199-207.
- Bühlmann, H. and Straub, E. (1970). Glaubwürdigkeit für Schadensätze.
  *Bulletin of the Swiss Association of Actuaries*, 70, 111-133.
- Clark, D. R. (2003). LDF curve-fitting and stochastic reserving:
  A maximum likelihood approach. *CAS Forum*, Fall 2003.
- Mack, T. (1993). Distribution-free calculation of the standard error
  of chain ladder reserve estimates. *ASTIN Bulletin*, 23(2), 213-225.
- Sherman, R. E. (1984). Extrapolating, smoothing, and interpolating
  development factors. *Proceedings of the Casualty Actuarial Society*,
  71, 122-155.
- Stanard, J. N. (1985). A simulation test of prediction errors of loss
  reserve estimation techniques. *Proceedings of the Casualty Actuarial
  Society*, 72, 124-153.
</content>
</invoke>
