---
title: "Projection: SA, ED, CL methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projection: SA, ED, CL methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

`fit_ratio()` projects cumulative loss ratio per cohort from a
`Triangle` object. Three methods are available; this vignette
explains the trade-offs.

## Notation

For cohort $i$ at dev $k$:

- $C^L_{i,k}$ — cumulative loss
- $C^P_{i,k}$ — cumulative risk premium (exposure)
- $f_k = C^L_{k+1} / C^L_k$ — age-to-age (chain ladder) factor
- $g_k = \Delta C^L_k / C^P_k$ — exposure-driven intensity
- maturity point $m_g$ — dev at which $f_k$ stabilises
  for group $g$ (detected from CV / RSE thresholds)

## Method 1: Stage-Adaptive (`"sa"`, default)

The default method exploits the fact that $f_k$ is volatile early
and stable late, while $g_k$ behaves the opposite way. SA switches
estimators at the maturity point:

$$
\hat{C}^L_{i,k+1} \;=\;
\begin{cases}
\hat{C}^L_{i,k} + g_k \cdot C^P_{i,k} & k < m_g \quad \text{(ED before maturity)} \\
f_k \cdot \hat{C}^L_{i,k}              & k \ge m_g \quad \text{(CL after maturity)}
\end{cases}
$$

Behaviour:

- **Before maturity**: anchors the loss estimate to premium volume.
  Avoids the volatile-link explosion that classical CL suffers when
  early $f_k$ are noisy.
- **After maturity**: preserves the cohort's own observed level.
  Avoids the "all cohorts converge to the average" behaviour that
  pure ED suffers in the tail.

When to use:

- Long-tail products where development extends across many years.
- Recent cohorts (immature data) mixed with older cohorts (matured).
- Health insurance cohorts with structural pre-/post-maturity
  difference (e.g. waiting period transitions).

```{r}
library(lossratio)
data(experience)
tri <- as_triangle(
  experience[coverage == "surgery"],
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  exposure = "incr_exposure"
)

ratio_sa <- fit_ratio(tri, method = "sa")        # default
plot(ratio_sa, metric = "ratio")
summary(ratio_sa)
```

## Method 2: Exposure-Driven (`"ed"`)

All future increments use ED:

$$
\hat{C}^L_{i,k+1} = \hat{C}^L_{i,k} + g_k \cdot C^P_{i,k}
$$

Behaviour:

- Stable when premium volume is informative across full development.
- Loses the cohort-specific level signal — cohorts with higher
  observed loss converge toward the group-level $g_k$.

When to use:

- Short-tail products where chain ladder offers no advantage.
- Sparse data where age-to-age factors are unreliable across all
  links.
- Comparing against SA / CL for sanity check.

```{r}
ratio_ed <- fit_ratio(tri, method = "ed")
plot(ratio_ed, metric = "ratio")
```

## Method 3: Classical Chain Ladder (`"cl"`)

Classical Mack (1993) model:

$$
\hat{C}^L_{i,k+1} = f_k \cdot \hat{C}^L_{i,k}
$$

Behaviour:

- Standard reserving practice. Equivalent to `fit_cl(tri, loss = "loss")`
  for the loss projection, but `fit_ratio()` additionally projects
  exposure forward via CL on `exposure` and computes the loss-ratio
  uncertainty via the delta method.
- Volatile when early $f_k$ are noisy — small denominators amplify
  link errors.

When to use:

- Mature, stable portfolios where age-to-age factors are
  well-behaved across the full development.
- Reserving exercises where regulators expect the classical Mack
  form for documentation.

```{r}
ratio_cl <- fit_ratio(tri, method = "cl")
plot(ratio_cl, metric = "ratio")
```

## Comparison

```{r}
ratios <- list(
  sa = fit_ratio(tri, method = "sa"),
  ed = fit_ratio(tri, method = "ed"),
  cl = fit_ratio(tri, method = "cl")
)

# Cohort-level summary
summary(ratios$sa)$loss_ult
summary(ratios$ed)$loss_ult
summary(ratios$cl)$loss_ult
```

## Variance and confidence intervals

`fit_ratio()` reports analytical standard errors for `L/P`. Two variants:

- `se_method = "fixed"` (default) — premium treated as fixed,
  $\mathrm{SE}(L/P) \approx \mathrm{SE}(L)/P$. Strictly the delta method
  with `Var(P) = 0` and `Cov(L,P) = 0`.
- `se_method = "delta"` — full delta method including premium uncertainty
  and loss-premium correlation `rho`:

$$
\mathrm{Var}(L/P) \approx \frac{\mathrm{Var}(L)}{P^2}
  + \frac{L^2 \mathrm{Var}(P)}{P^4}
  - \frac{2 \rho L \mathrm{SE}(L) \mathrm{SE}(P)}{P^3}
$$

Bootstrap intervals are also available:

```{r}
ratio_boot <- fit_ratio(tri, method = "sa", bootstrap = TRUE, B = 1000, seed = 1)
summary(ratio_boot)
```

## Choosing a method

SA combines ED before the maturity point with CL after, so `"sa"` is
the natural default. `"cl"` and `"ed"` are special cases that apply
only when one of SA's two regions becomes redundant.

```
Default is "sa" — ED before maturity, CL after.

Pick "cl" or "ed" only as special cases:
  ├── All cohorts are already past maturity
  │     → "cl"  (no ED region, so SA reduces to CL)
  └── Loss development is unstable across all dev and exposure is
      the more reliable signal
        → "ed"  (CL region is better served by exposure)
```

In practice: **start with `"sa"`** (the default), then run `"cl"`
and `"ed"` for sensitivity. If all three agree, the projection is
robust. If they diverge, inspect maturity detection and the
underlying ATA factors.

## Maturity input

For `method = "sa"` the maturity point $m_g$ determines where the
projection switches from ED to CL. By default `fit_ratio()` /
`fit_loss()` use `maturity = "auto"`, which calls
`detect_maturity()` internally with default thresholds. Other forms:

- `maturity = NULL` — disable detection (only meaningful for
  non-SA methods or for `fit_ata()` standalone).
- `maturity = maturity_spec(max_cv = 0.05, min_run = 2L)` — lazy
  spec forwarding custom thresholds to `detect_maturity()` at fit
  time (leakage-safe for `backtest()`).
- `maturity = detect_maturity(tri, ...)` — pre-built `Maturity`
  object, fixed across refits.
- `maturity = maturity_at(coverage = "surgery", change = 4)` — manual
  per-group override (e.g. company-standard $k^*$).
