---
title: "Triangle and Link: data structures and factor diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Triangle and Link: data structures and factor diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

Before fitting a chain ladder or loss-ratio model, it pays to inspect
the underlying triangle and the per-link factor table derived from it.
This vignette covers the `Triangle` and `Link` data structures and
their diagnostic plots; for maturity detection see
`vignette("maturity")`.

## Triangle-level diagnostics

For brevity this vignette uses the `SUR` group only — every step
generalises to multi-group input.

```{r}
library(lossratio)
data(experience)
exp <- as_experience(experience)[cv_nm == "SUR"]
tri <- build_triangle(exp, group_var = cv_nm)
```

### Cohort trajectories

```{r}
plot(tri)                              # cumulative loss-ratio trajectories per cohort
plot(tri, value_var = "lr")            # incremental loss ratio instead of clr
plot(tri, summary = TRUE)              # raw + overlay (mean / median / weighted)
```

The `summary = TRUE` overlay computes mean, median, and weighted
clr at each dev and overlays them on the cohort lines.
Useful for spotting cohorts that deviate from the central tendency.

### Cell heatmap

```{r}
plot_triangle(tri)                            # clr in each cell
plot_triangle(tri, value_var = "lr")          # incremental loss ratio

# detail labels (ratio + loss/rp amounts) are 2-line — use quarterly cells
tri_q <- build_triangle(exp, group_var = cv_nm,
                        cohort_var = "uyq", dev_var = "elap_q")
plot_triangle(tri_q, label_style = "detail")  # ratio + (loss / rp) amounts
```

### Group statistics by dev

```{r}
sm <- summary(tri)
head(sm)
```

Returns a `TriangleSummary` object with mean / median / weighted
loss ratios per (group, dev) cell.

## Link / factor diagnostics

The `Link` object is the link table (age-to-age factor table) built
from the triangle. In single-variable mode it carries the observed
ATA factors; with `exposure_var` it carries the ED-style intensities
$g_k = \Delta C^L_k / C^P_k$.

```{r}
ata <- build_link(tri, value_var = "closs")
sm  <- summary(ata, model = "ata", alpha = 1)
head(sm)
```

The `summary()` method on a `Link` object (single-variable mode)
computes per-link statistics that drive maturity detection:

- `mean`, `median`, `wt` — descriptive averages of observed ata
  factors at each link (excluding cohorts where the link is not
  observed).
- `cv` — coefficient of variation of the observed factors (relative
  spread, alpha-independent).
- `f` — WLS-estimated factor (volume-weighted by `value_from^alpha`).
- `f_se`, `rse` — WLS standard error and relative standard error.
- `sigma` — Mack residual sigma per link.
- `n_obs`, `n_valid`, `n_inf`, `n_nan`, `valid_ratio` — observation
  counts and the share of finite ATA factors per link.

### Diagnostic plots for the link table

```{r}
plot(ata, type = "cv")            # CV vs ata link with maturity overlay
plot(ata, type = "rse")           # RSE vs ata link
plot(ata, type = "summary")       # mean / median / wt overlay per link
plot(ata, type = "box")           # boxplot of observed ata per link
plot(ata, type = "point")         # scatter of observed ata per link
```

### Triangle of ATA factors

```{r}
la <- list(size = 2.5)                            # shrink labels
plot_triangle(ata, label_args = la)               # heatmap of observed factors
plot_triangle(ata, label_args = la, show_maturity = TRUE)    # overlay maturity line

# detail labels are two lines and overlap on monthly cells — rebuild on the
# quarterly triangle so the labels fit
ata_q <- build_link(tri_q, value_var = "closs")
plot_triangle(ata_q, label_style = "detail")      # factor + (loss / rp) amounts
```

The heatmap colours each cell by `log(ata / median(ata))` within its
link, so column-wise colour distinguishes cohorts that deviate from
the link's median.

### ED diagnostics

```{r}
ed <- build_link(tri, value_var = "closs", exposure_var = "crp")
sm <- summary(ed, model = "ed", alpha = 1)
head(sm)

plot(ed, type = "summary")
plot(ed, type = "box")
plot_triangle(ed, label_args = la)
```

`summary(link, model = "ed")` is the ED-side analogue of the
single-variable `summary()`, computing per-link statistics for the
intensity $g_k = \Delta C^L_k / C^P_k$.

## Validation before building

If gaps in the development sequence are suspected, inspect them before
`build_triangle()`:

```{r}
gaps <- validate_triangle(exp, group_var = cv_nm,
                          cohort_var = "uym", dev_var = "elap_m")
head(gaps)
```

Returns a `TriangleValidation` object with one row per cohort that
has non-consecutive development periods. An empty result means the
triangle is clean.

If gaps exist, options:

- Fix the data source (preferred).
- Drop offending cohorts.
- Pass `fill_gaps = TRUE` to `build_triangle()` to zero-fill missing
  cells (use with care — inflates `n_obs`).

## Recent-diagonal subset

When older cohorts are no longer representative (rate change,
reserving regime shift), restrict estimation to the recent calendar
diagonals:

```{r}
fit_ata(tri, value_var = "closs", alpha = 1, recent = 12)  # last 12 calendar diagonals
fit_cl(tri, value_var = "closs", recent = 12)
fit_lr(tri, recent = 12)
```

`recent = K` keeps only rows whose calendar position
(`rank(cohort) + dev - 1`) is among the latest `K` per group.

## Workflow checklist

Before fitting:

1. `validate_triangle()` — schema and gap check.
2. `build_triangle()` — canonical shape with derived columns.
3. `plot(tri)` / `plot_triangle(tri)` — visual inspection.
4. `summary(tri)` — group-level central tendency.
5. `build_link()` + `plot(link, type = "cv")` — link stability.
6. `detect_maturity()` — verify maturity detection produces a
   sensible point per group (see `vignette("maturity")`).
7. `detect_regime()` (optional) — structural change diagnosis.

Then fit `fit_lr()` / `fit_cl()` with confidence in the input data.

## See also

- `vignette("getting-started")` — full pipeline overview.
- `vignette("maturity")` — maturity detection from the link table.
- `vignette("regime")` — `detect_regime()` deep dive.
- `vignette("projection")` — projection method choice.
