---
title: "Triangle, Link and Maturity: data structures and factor diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Triangle, Link and Maturity: data structures and factor diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

Before fitting a chain ladder or loss-ratio model, it pays to inspect
the underlying triangle and the per-link factor table derived from it.
This vignette covers the `Triangle` and `Link` data structures, their
diagnostic plots, and `detect_maturity()` — the dev-axis test that
determines where the link table's ATA factors are stable enough to
trust for chain-ladder projection.

## Triangle-level diagnostics

For brevity this vignette uses the `surgery` group only — every step
generalises to multi-group input.

```{r}
library(lossratio)
data(experience)
exp <- experience[coverage == "surgery"]
tri <- as_triangle(
  exp,
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_premium"
)
```

### Cohort trajectories

```{r}
plot(tri)                              # cumulative loss-ratio trajectories per cohort
plot(tri, metric = "incr_ratio")       # incremental loss ratio instead of cumulative
plot(tri, summary = TRUE)              # raw + overlay (mean / median / weighted)
```

The `summary = TRUE` overlay computes mean, median, and weighted
ratio at each dev and overlays them on the cohort lines.
Useful for spotting cohorts that deviate from the central tendency.

### Cell heatmap

```{r}
plot_triangle(tri, metric = "ratio")          # cumulative ratio
plot_triangle(tri, metric = "incr_ratio")     # incremental ratio

# detail labels (ratio + loss/premium amounts) are 2-line — use quarterly cells
tri_q <- as_triangle(exp, groups = "coverage", cohort = "uy_m", calendar = "cy_m", loss = "incr_loss", premium = "incr_premium", grain = "Q")
plot_triangle(tri_q, label_style = "detail") # ratio + (loss / premium)
```

### Group statistics by dev

```{r}
sm <- summary(tri)
head(sm)
```

Returns a `TriangleSummary` object with mean / median / weighted
loss ratios per (group, dev) cell.

## Link / factor diagnostics

The `Link` object is the link table (age-to-age factor table) built
from the triangle. In single-variable mode it carries the observed
ATA factors; with `premium` it carries the ED-style intensities
$g_k = \Delta C^L_k / C^P_k$.

```{r}
ata <- as_link(tri, loss = "loss")
sm  <- summary(ata, model = "ata", alpha = 1)
head(sm)
```

The `summary()` method on a `Link` object (single-variable mode)
computes per-link statistics that drive maturity detection:

- `mean`, `median`, `wt` — descriptive averages of observed ata
  factors at each link (excluding cohorts where the link is not
  observed).
- `cv` — coefficient of variation of the observed factors (relative
  spread, alpha-independent).
- `f` — WLS-estimated factor (volume-weighted by `loss_from^alpha`).
- `f_se`, `rse` — WLS standard error and relative standard error.
- `sigma` — Mack residual sigma per link.
- `n_cohorts`, `n_valid`, `n_inf`, `n_nan`, `valid_ratio` — observation
  counts and the share of finite ATA factors per link.

### Diagnostic plots for the link table

```{r}
plot(ata, type = "cv")            # CV vs ata link with maturity overlay
plot(ata, type = "rse")           # RSE vs ata link
plot(ata, type = "summary")       # mean / median / wt overlay per link
plot(ata, type = "box")           # boxplot of observed ata per link
plot(ata, type = "point")         # scatter of observed ata per link
```

### Triangle of ATA factors

```{r}
plot_triangle(ata, label_size = 2.5)              # heatmap of observed factors
plot_triangle(ata, label_size = 2.5, show_maturity = TRUE)   # overlay maturity line

# detail labels are two lines and overlap on monthly cells — rebuild on
# the quarterly triangle so the two-line "factor (loss / premium)" text
# has room (label_size auto-shrinks to 2.2 in detail mode).
ata_q <- as_link(tri_q, loss = "loss")
plot_triangle(ata_q, label_style = "detail")      # factor + (loss / premium)
```

The heatmap colours each cell by `log(ata / median(ata))` within its
link, so column-wise colour distinguishes cohorts that deviate from
the link's median.

### ED diagnostics

```{r}
ed <- as_link(tri, loss = "loss", exposure = "premium")
sm <- summary(ed, model = "ed", alpha = 1)
head(sm)

plot(ed, type = "summary")
plot(ed, type = "box")
plot_triangle(ed)                                  # ED triangle (default size)
```

`summary(link, model = "ed")` is the ED-side analogue of the
single-variable `summary()`, computing per-link statistics for the
intensity $g_k = \Delta C^L_k / C^P_k$.

## Maturity detection

The maturity point is the development link beyond which age-to-age
factors are stable enough to trust for chain-ladder projection. Used
internally by `fit_ratio(method = "sa")` to switch from ED to CL.

### Detecting maturity

`detect_maturity()` takes a `Triangle` directly — the underlying
single-variable `Link` and its WLS summary are built internally:

```{r}
mat <- detect_maturity(
  tri,
  loss            = "loss",
  max_cv          = 0.15,    # CV must be below this
  max_rse         = 0.05,    # RSE must be below this
  min_valid_ratio = 0.5,     # at least 50% finite cohorts at the link
  min_n_valid     = 3L,      # at least 3 finite cohorts
  min_run         = 1L       # at least 1 consecutive mature link
)

print(mat)
```

A row per group with the first development link satisfying all
thresholds, carrying the link's full statistics. The threshold
arguments are also stored as attributes on the returned `Maturity`
object.

### Threshold semantics

- `max_cv` — coefficient of variation of the observed ATA
  factors at the link. Caps relative spread regardless of `alpha`.
- `max_rse` — relative standard error of the WLS-estimated
  factor `f`. Captures parameter uncertainty rather than residual
  spread.
- `min_valid_ratio` — minimum share of cohorts with a finite ATA at
  the link. Guards against links where most observations are zero
  / NA / Inf.
- `min_n_valid` — minimum count of finite cohorts at the link.
  Absolute floor for thin-data tails.
- `min_run` — minimum number of *consecutive* mature links. With
  `min_run = 1L` (default) the first qualifying link wins; setting
  it to `2L` or higher requires sustained stability.

Tune these to your portfolio's volatility profile. Tight thresholds
(e.g. `max_cv = 0.05`) push maturity later; loose thresholds
push it earlier.

### Use in fitting

`detect_maturity()` is also called internally by `fit_ata()` and
`fit_cl()` via the `maturity` argument. Pass `maturity_spec()` to
forward custom detection thresholds, or `"auto"` for defaults
(`fit_ratio(method = "sa")` and `fit_loss()` use `"auto"` by default):

```{r, eval = FALSE}
fit_ata(tri, loss = "loss",
        maturity = maturity_spec(max_cv = 0.08, min_run = 2L))

fit_cl(tri, loss = "loss",
       maturity = maturity_spec(max_cv = 0.08))

fit_ratio(tri, method = "sa",
          maturity = maturity_spec(max_cv = 0.08))
```

The `maturity` argument accepts four forms: `NULL` (no detection),
a pre-built `Maturity` object (e.g. from `detect_maturity()` or
`maturity_at()` for manual override), the string `"auto"`
(internal `detect_maturity()` with defaults), or a function of one
triangle argument returning a `Maturity` (typically built with
`maturity_spec()`).

For `fit_ratio(method = "sa")` the detected maturity point determines
the dev at which the projection switches from ED (early dev) to CL
(later dev).

### Group-wise output

For multi-group triangles `detect_maturity()` returns one row per
group, each detected independently with the same thresholds:

```{r}
tri_all <- as_triangle(
  experience,
  groups   = "coverage",
  cohort   = "uy_m",
  calendar = "cy_m",
  loss     = "incr_loss",
  premium  = "incr_premium"
)
detect_maturity(tri_all, loss = "loss")
```

The link diagnostic plot makes the result easier to read — each cv
gets its own panel and the vertical line marks the maturity point
where CV first drops below `max_cv`:

```{r}
plot(as_link(tri_all, loss = "loss"), type = "cv")
```

## Validation before building

If gaps in the development sequence are suspected, inspect them before
`as_triangle()`:

```{r}
gaps <- validate_triangle(exp, groups = "coverage", cohort = "uy_m", dev = "dev_m")
head(gaps)
```

Returns a `TriangleValidation` object with one row per cohort that
has non-consecutive development periods. An empty result means the
triangle is clean.

If gaps exist, options:

- Fix the data source (preferred).
- Drop offending cohorts.
- Pass `fill_gaps = TRUE` to `as_triangle()` to zero-fill missing
  cells (use with care — inflates `n_cohorts`).

## Recent-diagonal subset

When older cohorts are no longer representative (rate change,
reserving regime shift), restrict estimation to the recent calendar
diagonals:

```{r}
fit_ata(tri, loss = "loss", alpha = 1, recent = 12)  # last 12 calendar diagonals
fit_cl(tri, loss = "loss", recent = 12)
fit_ratio(tri, recent = 12)
```

`recent = K` keeps only rows whose calendar position
(`rank(cohort) + dev - 1`) is among the latest `K` per group.

## Workflow checklist

Before fitting:

1. `validate_triangle()` — schema and gap check.
2. `as_triangle()` — canonical shape with derived columns.
3. `plot(tri)` / `plot_triangle(tri)` — visual inspection.
4. `summary(tri)` — group-level central tendency.
5. `as_link()` + `plot(link, type = "cv")` — link stability.
6. `detect_maturity()` — verify maturity detection produces a
   sensible point per group.
7. `detect_regime()` (optional) — structural change diagnosis.

Then fit `fit_ratio()` / `fit_cl()` with confidence in the input data.

## See also

- `vignette("getting-started")` — full pipeline overview.
- `vignette("regime")` — `detect_regime()` deep dive.
- `vignette("projection")` — projection method choice.
